---
title: "Core data exploration"
## format: html
---

```{r}
library(edgeR)
library(SingleCellExperiment)
library(data.table)
library(ggplot2)
library(dplyr)
library(here)
ddir <- here("processed-data")
```


```{r ERC DGE results}
## fpdge <- file.path(ddir, "08_pseudoBulkDGE_sn/01_pseudobulk_data_sn", "sce_pseudo_DGE-cell_type_anno.RDS")
## pdge <- readRDS(fpdge)
fpdge <- file.path(ddir, "13_compile_DGE/01_compile_DGE/sn_fine", "DGE_results_carrier_sn_fine.Rds")
pdge <- readRDS(fpdge)
sumdt <- fread(file.path(ddir, "13_compile_DGE/01_compile_DGE/sn_fine", "vlmf_model_summary_sn_fine.csv"))
```

```{r-console}
> head(sumdt)
      V1 cell_type_broad cluster     mod n_genes n_FDR05   nUP nDown
   <int>          <char>  <char>  <char>   <int>   <int> <int> <int>
1:     1           Astro Astro.1 carrier   16867       7     5     2
2:     2           Astro Astro.2 carrier    9994      58    10    48
3:     3           Astro Astro.3 carrier   11282      23     5    18
4:     4           Astro Astro.4 carrier    8135       0     0     0
5:     5           Astro Astro.5 carrier    1793       0     0     0
6:     6           Oligo Oligo.1 carrier   13881       0     0     0

> head(pdge)
# A tibble: 6 × 21
  data_type cluster contrast seqnames    start      end width strand source type  gene_id         gene_version gene_name       gene_type      vlmf_logFC vlmf_AveExpr vlmf_t vlmf_P.Value vlmf_adj.P.Val vlmf_B cell_type_broad
  <chr>     <fct>   <chr>    <fct>       <int>    <int> <int> <fct>  <fct>  <fct> <chr>           <chr>        <chr>           <chr>               <dbl>        <dbl>  <dbl>        <dbl>          <dbl>  <dbl> <fct>
1 sn_fine   Astro.1 carrier  chr15    82339993 82349475  9483 -      HAVANA gene  ENSG00000278662 6            GOLGA6L10       protein_coding     -1.55          1.83  -6.46  0.000000659         0.0111   5.60 Astro
2 sn_fine   Astro.1 carrier  chr22    38818452 38844028 25577 -      HAVANA gene  ENSG00000221890 6            NPTXR           protein_coding      0.751         4.29   5.73  0.00000436          0.0310   4.28 Astro
3 sn_fine   Astro.1 carrier  chr1     99264292 99309590 45299 +      HAVANA gene  ENSG00000117600 14           PLPPR4          protein_coding      1.50          2.27   5.64  0.00000552          0.0310   3.76 Astro
4 sn_fine   Astro.1 carrier  chr6     40344346 40387194 42849 -      HAVANA gene  ENSG00000226070 2            LINC00951       lncRNA              1.19          1.81   5.47  0.00000885          0.0373   3.21 Astro
5 sn_fine   Astro.1 carrier  chr20     1266280  1309328 43049 +      HAVANA gene  ENSG00000101298 15           SNPH            protein_coding      0.538         4.31   5.29  0.0000140           0.0404   3.17 Astro
6 sn_fine   Astro.1 carrier  chr10    47538377 47551954 13578 -      HAVANA gene  ENSG00000290453 1            ENSG00000290453 lncRNA             -0.911         3.66  -5.29  0.0000144           0.0404   3.16 Astro

 ```

```{r the 88 Astro genes}
stopifnot(unique(pdge$data_type)=='sn_fine')
## 88 Astro DEGs FDR<0.05
degs <- as.data.table(pdge)[grepl("^Astro", cluster) & vlmf_adj.P.Val < 0.05 & contrast=='carrier',
   .(cluster, gene_id, gene_name, gene_type, vlmf_logFC, vlmf_AveExpr, vlmf_t, vlmf_P.Value, vlmf_adj.P.Val, cell_type_broad)]
## nrow(degs) # 88
#length(unique(degs$gene_name)) # 86
degs[gene_name %in% degs[duplicated(degs$gene_name)]$gene_name, .(gene_name, gene_id, cluster)]
## only these two genes are shared between clusters
#1:    GABBR2 ENSG00000136928 Astro.1
#2:    ZNF536 ENSG00000198597 Astro.2
#3:    GABBR2 ENSG00000136928 Astro.3
#4:    ZNF536 ENSG00000198597 Astro.3

### Show the same for Oligo.3:
odegs <- as.data.table(pdge)[grepl("^Oligo", cluster) & vlmf_adj.P.Val < 0.05 & contrast=='carrier',
   .(cluster, gene_id, gene_name, gene_type, vlmf_logFC, vlmf_AveExpr, vlmf_t, vlmf_P.Value, vlmf_adj.P.Val, cell_type_broad)]

```


```{r get expression data for these genes per brain and cell type}
## pseudobulk data

pb_fn <- file.path(ddir,  "08_pseudoBulkDGE_sn/01_pseudobulk_data_sn","sce_pseudo_DGE-cell_type_anno.RDS")
pb <- readRDS(pb_fn)
## Drop donor Br1289
pb <- pb[, pb$BrNum != 'Br1289']

astro_cls <- unique(as.character(degs$cluster))
pbdex <- pb[degs$gene_id, pb$registration_variable %in% astro_cls]
pbdex$cluster <- as.character(pbdex$registration_variable)
```

```{r-console}
## corner of the logcounts assay
> hh(assays(pbdex)$logcounts)
                Br1039_Astro.1 Br1556_Astro.1 Br1691_Astro.1 Br1706_Astro.1 Br2305_Astro.1 Br2582_Astro.1 Br5161_Astro.1 Br5212_Astro.1
ENSG00000278662       2.610157       2.831021      0.5297312     0.05268869       1.855498       2.683376       1.330184       3.277047
ENSG00000221890       4.114118       4.041462      5.5256579     4.79502073       4.280823       4.042058       4.752787       4.247719
ENSG00000117600       1.837585       1.257096      3.0546350     3.55175869       2.809136       1.828722       3.689752       2.699462
ENSG00000226070       1.976645       1.805763      2.5898594     2.67402766       2.177195       1.503035       2.627911       2.564164
ENSG00000101298       4.922988       4.425846      5.5610530     4.77638361       4.346402       3.910505       5.020642       4.699048
ENSG00000290453       3.955009       4.923233      2.2857618     2.67402766       3.876996       4.519407       4.102840       4.963852

> table(pbdex$cluster)

Astro.1 Astro.2 Astro.3
     29      30      30

> rowRanges(pbdex)
GRanges object with 88 ranges and 6 metadata columns:
                  seqnames              ranges strand |   source     type         gene_id gene_version   gene_name      gene_type
                     <Rle>           <IRanges>  <Rle> | <factor> <factor>     <character>  <character> <character>    <character>
  ENSG00000278662    chr15   82339993-82349475      - |   HAVANA     gene ENSG00000278662            6   GOLGA6L10 protein_coding
  ENSG00000221890    chr22   38818452-38844028      - |   HAVANA     gene ENSG00000221890            6       NPTXR protein_coding
  ENSG00000117600     chr1   99264292-99309590      + |   HAVANA     gene ENSG00000117600           14      PLPPR4 protein_coding
  ENSG00000226070     chr6   40344346-40387194      - |   HAVANA     gene ENSG00000226070            2   LINC00951         lncRNA
  ENSG00000101298    chr20     1266280-1309328      + |   HAVANA     gene ENSG00000101298           15        SNPH protein_coding

```


### Comparison boxplots BEFORE vs AFTER mediation

LEFT panel (Baseline): E2 vs E4 boxplots showing Oligo.3 gene expression before adding the mediator (Astro PLP1) as a covariate. The expression values are "cleaned" using jaffelab::cleaningY to remove effects of nuisance covariates (Sex, Age, Anc_Afr, pseudo_expr_chrM_ratio) while preserving the APOE effect.

RIGHT panel (After mediation): E2 vs E4 boxplots showing the same outcome gene expression, but now cleaned to remove both nuisance covariates AND the mediator gene expression. If the mediator explains part of the APOE→Oligo.3 effect, the E2 vs E4 difference should shrink.

How cleaningY(y, mod, P) works:
Fits y = mod %*% beta (full model)
Returns y - (effects from columns P+1 to end)
With P=4, preserves APOE columns (1-4), removes covariate effects
Input data needed:
Outcome expression matrix (y_logcpm): logCPM from edgeR::cpm(dge_base, log=TRUE) for Oligo.3
Mediator expression vector: From med_expr[med_gene_id, ] aligned to outcome samples
Design matrices:
Baseline: ~0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio
Mediation: same + med_vec
Sample metadata: dge_base$samples$APOE_syn to group E2.E2 vs E3.E4/E4.E4
Verified calculation (FOS gene + PLP1 mediator):
Baseline E4-E2 difference: 1.995 logCPM
After PLP1 mediation: 1.744 logCPM
12.6% reduction in APOE effect


```{r erc_astro plots}
library(qs2)
library(ggplot2)
library(tidyr)

## assumes 01_mediation_screen.R was already run for this scenario
## load med_degs and med_expr for out-erc_astro

cdir <- here("code")
out_dt <- file.path(cdir, "22_Mediation", "out-erc_astro")
fn_med_degs <- file.path(out_dt, "med_degs.qs2")
fn_med_expr <- file.path(out_dt, "med_expr.qs2")
med_degs <- qs_read(fn_med_degs)
med_expr <- qs_read(fn_med_expr)
baseline_dir <- file.path(out_dt, "baseline")
fn_dge_base <- file.path(out_dt, "dge_base.qs2")
dge_base <- qs_read(fn_dge_base)



single_baseline_key="80c146156f90daa41a2eef910b7d1c83"
baseline_file <- file.path(baseline_dir, sprintf("baseline_%s.tab.gz", single_baseline_key))
baseline_results_dt <- data.table::fread(baseline_file)


APOE_carrier_colors_dark <- c(`E2+`="#3B918B", `E4+`="#B7522A")
APOE_carrier_colors <- c(`E2+`="#51B8B1", `E4+`="#D97D59")

plotMedComp <- function(outcome_gene,
                           mediator_gene,
                           dge_base = NULL,
                           med_degs = NULL,
                           med_expr = NULL,
                           n_apoe_cols = 4) {
    # Check for required objects in parent environment if not provided
    if (is.null(dge_base)) {
        if (!exists("dge_base", envir = parent.frame())) {
            stop("dge_base must be provided or exist in the calling environment.")
        }
        dge_base <- get("dge_base", envir = parent.frame())
    }
    if (is.null(med_degs)) {
        if (!exists("med_degs", envir = parent.frame())) {
            stop("med_degs must be provided or exist in the calling environment.")
        }
        med_degs <- get("med_degs", envir = parent.frame())
    }
    if (is.null(med_expr)) {
        if (!exists("med_expr", envir = parent.frame())) {
            stop("med_expr must be provided or exist in the calling environment.")
        }
        med_expr <- get("med_expr", envir = parent.frame())
    }

    baseline_formula_str <- "0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio"
    mediation_formula_str <- paste(baseline_formula_str, "+ med_vec")
    baseline_formula <- as.formula(paste("~", baseline_formula_str))
    mediation_formula <- as.formula(paste("~", mediation_formula_str))

    if (!exists("baseline_results_dt", envir = parent.frame())) {
        stop("baseline_results_dt must be loaded in the environment.")
    }
    baseline_results_dt <- get("baseline_results_dt", envir = parent.frame())

    if (!exists("out_dt", envir = parent.frame())) {
        stop("out_dt must be loaded in the environment.")
    }
    out_dt <- get("out_dt", envir = parent.frame())

    mediator_cluster <- NULL
    mediator_gene_name <- mediator_gene
    if (grepl(":", mediator_gene, fixed = TRUE)) {
        parts <- strsplit(mediator_gene, ":", fixed = TRUE)[[1]]
        mediator_cluster <- parts[1]
        mediator_gene_name <- parts[2]
    }

    med_rows <- med_degs[gene_name == mediator_gene_name]
    if (!is.null(mediator_cluster)) {
        med_rows <- med_rows[cluster == mediator_cluster]
    }
    if (nrow(med_rows) == 0) {
        stop("Mediator gene not found: ", mediator_gene)
    }
    if (nrow(med_rows) > 1) {
        warning(sprintf(
            "Multiple matches for mediator gene '%s' found in clusters: %s. Using last match; use 'cluster:gene' to disambiguate.",
            mediator_gene,
            paste(unique(med_rows$cluster), collapse = ", ")
        ))
    }
    med_row <- med_rows[nrow(med_rows)]

    run_label <- sprintf("%s|%s|%s", med_row$cluster, med_row$gene_id, med_row$gene_name)
    file_label <- gsub("\\|", "_", run_label)
    mediation_file <- file.path(out_dt, sprintf("%02d_%s.tab.gz", med_row$iter, file_label))
    mediation_results_dt <- data.table::fread(mediation_file)

    # 1. Get mediator info
    med_gid <- med_row$med_gene_id
    med_cluster <- as.character(med_row$cluster)

    # 2. Resolve outcome gene name -> gene_id
    if (is.null(dge_base$genes) || !("gene_name" %in% colnames(dge_base$genes))) {
        stop("dge_base$genes must contain a gene_name column.")
    }
    matches <- rownames(dge_base$genes)[dge_base$genes$gene_name == outcome_gene]
    if (length(matches) == 0) stop("Outcome gene not found in dge_base: ", outcome_gene)
    out_gid <- matches[1]
    out_gname <- outcome_gene

    # 3. Get outcome expression (logCPM)
    y_logcpm <- edgeR::cpm(dge_base, log = TRUE)
    y_mat <- matrix(y_logcpm[out_gid, ], nrow = 1)
    rownames(y_mat) <- out_gid
    colnames(y_mat) <- colnames(y_logcpm)

    # 4. Align mediator to outcome samples
    dge_brnum <- gsub("_Oligo\\.3$", "", colnames(dge_base))
    med_vec_aligned <- med_expr[med_gid, dge_brnum]

    # 5. Build design matrices
    baseline_mod <- model.matrix(baseline_formula, data = dge_base$samples)

    sample_df_med <- dge_base$samples
    sample_df_med$med_vec <- med_vec_aligned
    mediation_mod <- model.matrix(mediation_formula, data = sample_df_med)

    # 6. Apply cleaningY (preserve APOE, remove covariates)
    y_cleaned_baseline <- jaffelab::cleaningY(y_mat, baseline_mod, P = n_apoe_cols)
    y_cleaned_mediation <- jaffelab::cleaningY(y_mat, mediation_mod, P = n_apoe_cols)

    # 7. Create plot data frame
    apoe_status <- dge_base$samples$APOE_syn
    carrier_group <- ifelse(apoe_status %in% c("E2.E2", "E2.E3"), "E2+",
                        ifelse(apoe_status %in% c("E3.E4", "E4.E4"), "E4+", "other"))

    plot_df <- data.frame(
        BrNum = dge_brnum,
        carrier_group = carrier_group,
        baseline_cleaned = as.numeric(y_cleaned_baseline[1, ]),
        mediation_cleaned = as.numeric(y_cleaned_mediation[1, ])
    )

    plot_df <- plot_df[plot_df$carrier_group %in% c("E2+", "E4+"), ]

    model_labels <- c(
        sprintf("Baseline\n(no %s)", mediator_gene),
        sprintf("After %s\n(%s) adjustment", mediator_gene, med_cluster)
    )

    plot_long <- plot_df |>
        tidyr::pivot_longer(
            cols = c(baseline_cleaned, mediation_cleaned),
            names_to = "model",
            values_to = "expression"
        ) |>
        dplyr::mutate(
            model = factor(model,
                           levels = c("baseline_cleaned", "mediation_cleaned"),
                           labels = model_labels)
        )

    annotation_df <- NULL
    if (!is.null(baseline_results_dt) && !is.null(mediation_results_dt)) {
        base_row <- baseline_results_dt[outcome_gene_id == out_gid, , drop = FALSE]
        med_row <- mediation_results_dt[outcome_gene_id == out_gid, , drop = FALSE]
        base_row <- if (nrow(base_row) > 0) base_row[1, ] else NULL
        med_row <- if (nrow(med_row) > 0) med_row[1, ] else NULL

        fdr_base <- if (!is.null(base_row) && "adj.P.Val" %in% names(base_row)) base_row$adj.P.Val else NA_real_
        tvar_base <- if (!is.null(base_row) && "t" %in% names(base_row)) base_row$t else NA_real_
        fdr_med <- if (!is.null(med_row) && "adj.P.Val" %in% names(med_row)) med_row$adj.P.Val else NA_real_
        tvar_med <- if (!is.null(med_row) && "t" %in% names(med_row)) med_row$t else NA_real_

        y_max <- max(plot_long$expression, na.rm = TRUE)
        y_min <- min(plot_long$expression, na.rm = TRUE)
        y_fixed <- y_max + pmax(0.08 * (y_max - y_min), 0.2)

        # Format FDR values: use fixed decimal notation for values >= 1e-4
        format_fdr <- function(fdr) {
            if (is.na(fdr)) return("NA")
            if (fdr >= 1e-4) {
                sprintf("%.5f", fdr)
            } else {
                sprintf("%.2e", fdr)
            }
        }

        annotation_df <- dplyr::tibble(
            model = factor(model_labels, levels = model_labels),
            label = c(
                sprintf("FDR: %s\nt: %.2f", format_fdr(fdr_base), tvar_base),
                sprintf("FDR: %s\nt: %.2f", format_fdr(fdr_med), tvar_med)
            ),
            y = y_fixed
        )
    }

    ggplot(plot_long, aes(x = carrier_group, y = expression, fill = carrier_group,
                        color = carrier_group)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA) +
        geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
        facet_wrap(~model) +
        scale_fill_manual(values = APOE_carrier_colors) +
        scale_color_manual(values = APOE_carrier_colors_dark) +
        labs(
            title = sprintf("%s (%s) Expression: E2+ vs E4+ Carriers",
                            out_gname, out_gid),
            subtitle = sprintf("Mediator: %s expression from %s cluster",
                               mediator_gene, med_cluster),
            x = "APOE Carrier Status",
            y = "Covariate-adjusted Expression (logCPM)",
            fill = "Carrier"
        ) +
        {
            if (!is.null(annotation_df)) {
                ggplot2::geom_text(
                    data = annotation_df,
                    ggplot2::aes(x = 1.5, y = y, label = label),
                    inherit.aes = FALSE,
                    size = 3.5
                )
            }
        } +
        theme_bw(base_size = 14) +
        theme(
            legend.position = "none",
            strip.text = element_text(size = 12, face = "bold")
        ) +
        coord_cartesian(clip = "off")
}


p <- plotMedComp(outcome_gene = "FOS", mediator_gene = "PLP1")
print(p)
```

```{r plots plots plots}
dnr=c('PLP1', 'MAG', 'MAL', 'MBP', 'SOX10', 'OPALIN')
upr=c('FOS', 'TLR2', 'STAT1', 'STAT4')
alldegs=c(dnr, upr)
medgenes=c('PLP1', 'MBP', 'OPALIN')
pdf_dir <- out_dt
pdf_files <- file.path(pdf_dir, sprintf("mediator_%s.pdf", medgenes))

plot_one_mediator <- function(mediator_gene, outcome_genes, out_file) {
    grDevices::pdf(out_file, width = 11, height = 8.5, onefile = TRUE)
    on.exit(grDevices::dev.off(), add = TRUE)

    for (outcome_gene in outcome_genes) {
        p <- plotMedComp(outcome_gene = outcome_gene, mediator_gene = mediator_gene)
        print(p)
    }

    invisible(out_file)
}

pdf_out <- mapply(
    plot_one_mediator,
    mediator_gene = medgenes,
    out_file = pdf_files,
    MoreArgs = list(outcome_genes = alldegs),
    SIMPLIFY = FALSE
)

pdf_out

```