---
title: "Plot mediation screening results"
## format: html
---

```{r}
library(edgeR)
library(SingleCellExperiment)
library(data.table)
library(ggplot2)
library(dplyr)
library(here)
ddir <- here("processed-data")
```



```{r setup}
library(qs2)
library(ggplot2)
library(tidyr)
## assumes 01_mediation_screen.R was already run for this scenario
## load med_degs and med_expr for out-erc_astro
cdir <- here("code")
med_degs <- NULL
med_expr <-NULL
dge_base <- NULL
baseline_results_dt <- NULL
out_dt <- NULL
mediation_index_dt <- NULL
baseline_index_dt <- NULL

loadMedData <- function(out_dir) {
  out_dt <<- out_dir
  fn_med_degs <- file.path(out_dt, "med_degs.qs2")
  fn_med_expr <- file.path(out_dt, "med_expr.qs2")

    if (!file.exists(fn_med_degs)) stop("Missing `med_degs.qs2`: ", fn_med_degs)
    if (!file.exists(fn_med_expr)) stop("Missing `med_expr.qs2`: ", fn_med_expr)
    med_degs <<- qs_read(fn_med_degs)
    if (!is.data.frame(med_degs) || nrow(med_degs) == 0) {
        stop("`med_degs` is empty or not a data.frame from: ", fn_med_degs)
    }
    med_expr <<- qs_read(fn_med_expr)
    if (!is.matrix(med_expr) || nrow(med_expr) == 0) {
        stop("`med_expr` is empty or not a matrix from: ", fn_med_expr)
    }
  baseline_dir <- file.path(out_dt, "baseline")
  fn_dge_base <- file.path(out_dt, "dge_base.qs2")
    if (!file.exists(fn_dge_base)) stop("Missing `dge_base.qs2`: ", fn_dge_base)
    dge_base <<- qs_read(fn_dge_base)
  fn_med_idx <- file.path(out_dt, "mediation_run_index.tsv")
  fn_base_idx <- file.path(out_dt, "baseline_index.tsv")

  mediation_index_dt <<- if (file.exists(fn_med_idx)) {
    data.table::fread(fn_med_idx)
  } else NULL
  baseline_index_dt <<- if (file.exists(fn_base_idx)) {
    data.table::fread(fn_base_idx)
  } else NULL
  baseline_file <- NULL
  if (!is.null(baseline_index_dt) && nrow(baseline_index_dt) > 0 && "out_file" %in% names(baseline_index_dt)) {
    if ("n_donors" %in% names(baseline_index_dt)) {
      setorderv(baseline_index_dt, cols = c("n_donors"), order = c(-1L))
    }
    baseline_file <- baseline_index_dt$out_file[1]
  } else {
    baseline_file <- list.files(baseline_dir, pattern = "^baseline_[0-9a-f]+\\.tab\\.gz$", full.names = TRUE)[1]
  }
  if (is.na(baseline_file) || !nzchar(baseline_file) || !file.exists(baseline_file)) {
    stop("Could not resolve baseline file in: ", baseline_dir)
  }
  baseline_results_dt <<- data.table::fread(baseline_file)
    if (!is.data.frame(baseline_results_dt) || nrow(baseline_results_dt) == 0) {
        stop("`baseline_results_dt` is empty from: ", baseline_file)
    }
}


```

```{r load mediation data}
## loading ERC Astro mediation data
 loadMedData(file.path(cdir, "22_Mediation", "out-erc_astro"))
```


```{r plot function and test}
## out_dt also required in the environment !
APOE_carrier_colors_dark <- c(`E2+`="#3B918B", `E4+`="#B7522A")
APOE_carrier_colors <- c(`E2+`="#51B8B1", `E4+`="#D97D59")

resolve_mediator_row <- function(mediator_gene, med_degs) {
    mediator_cluster <- NULL
    mediator_gene_name <- mediator_gene
    if (grepl(":", mediator_gene, fixed = TRUE)) {
        parts <- strsplit(mediator_gene, ":", fixed = TRUE)[[1]]
        mediator_cluster <- parts[1]
        mediator_gene_name <- parts[2]
    }
    med_rows <- med_degs[gene_name == mediator_gene_name |
            gene_id == mediator_gene_name | med_gene_id == mediator_gene_name ]
    if (!is.null(mediator_cluster)) {
        med_rows <- med_rows[cluster == mediator_cluster]
    }
    if (nrow(med_rows) == 0) {
        candidates <- unique(med_degs$gene_name)
        close <- candidates[agrep(mediator_gene_name, candidates, max.distance = 0.2)]
        stop("Mediator not found: ", mediator_gene, "\n",
            "Tried gene_name / gene_id / med_gene_id.\n",
            "Close matches: ", paste(head(close, 6), collapse = ", "))
    }
    if (nrow(med_rows) > 1) {
        warning(sprintf(
            "Multiple matches for mediator gene '%s' in clusters: %s. Using last match; use 'cluster:gene' to disambiguate.",
            mediator_gene, paste(unique(med_rows$cluster), collapse = ", ")
        ))
    }
    med_rows[nrow(med_rows)]
}

plotMedComp <- function(outcome_gene,
                           mediator_gene,
                           dge_base = NULL,
                           med_degs = NULL,
                           med_expr = NULL, med_region="ERC", ## for some labeling
                           n_apoe_cols = 4) {
    ## check for required objects in parent environment if not provided
    if (is.null(dge_base)) {
        if (!exists("dge_base", envir = parent.frame())) {
            stop("dge_base must be provided or exist in the calling environment.")
        }
        dge_base <- get("dge_base", envir = parent.frame())
    }
    if (is.null(med_degs)) {
        if (!exists("med_degs", envir = parent.frame())) {
            stop("med_degs must be provided or exist in the calling environment.")
        }
        med_degs <- get("med_degs", envir = parent.frame())
    }
    if (is.null(med_expr)) {
        if (!exists("med_expr", envir = parent.frame())) {
            stop("med_expr must be provided or exist in the calling environment.")
        }
        med_expr <- get("med_expr", envir = parent.frame())
    }

    baseline_formula_str <- "0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio"
    med_formula_str <- paste(baseline_formula_str, "+ med_vec")
    baseline_formula <- as.formula(paste("~", baseline_formula_str))
    med_formula <- as.formula(paste("~", med_formula_str))

    if (!exists("baseline_results_dt", envir = parent.frame())) {
        stop("baseline_results_dt must be loaded in the environment.")
    }
    baseline_results_dt <- get("baseline_results_dt", envir = parent.frame())

    if (!exists("out_dt", envir = parent.frame())) {
        stop("out_dt must be loaded in the environment.")
    }
    out_dt <- get("out_dt", envir = parent.frame())
    med_idx <- if (exists("mediation_index_dt", envir = parent.frame())) {
        get("mediation_index_dt", envir = parent.frame())
    } else {
        NULL
    }

    med_row <- resolve_mediator_row(mediator_gene, med_degs)

    run_label <- sprintf("%s|%s|%s", med_row$cluster, med_row$gene_id, med_row$gene_name)
    file_label <- gsub("\\|", "_", run_label)
    med_iter <- as.integer(med_row$iter)
    med_file <- character(0)
    if (!is.null(med_idx) && nrow(med_idx) > 0 &&
        all(c("iter", "med_gene_id", "out_file") %in% names(med_idx))) {
        idx_row <- med_idx[iter == med_iter & med_gene_id == med_row$med_gene_id]
        if (nrow(idx_row) > 0) {
            med_file <- idx_row$out_file[1]
        }
    }
    if (length(med_file) < 1 || !file.exists(med_file)) {
        med_file <- file.path(out_dt, sprintf("%03d_%s.tab.gz", med_iter, file_label))
    }
    if (!file.exists(med_file)) {
        candidates <- list.files(out_dt, pattern = sprintf("_%s\\.tab\\.gz$", file_label), full.names = TRUE)
        if (length(candidates) > 1) {
            info <- file.info(candidates)
            candidates <- candidates[order(info$mtime, decreasing = TRUE)]
        }
        med_file <- candidates[1]
    }
    if (is.na(med_file) || !nzchar(med_file) || !file.exists(med_file)) {
        stop("Mediation results file not found for mediator: ", mediator_gene)
    }
    med_res_dt <- data.table::fread(med_file)

    ## get mediator info file
    med_gid <- med_row$med_gene_id
    med_cluster <- as.character(med_row$cluster)
    if (!med_gid %in% rownames(med_expr)) {
        stop("Mediator expression not found in `med_expr` for: ", med_gid)
    }

    ## resolve outcome gene name -> gene_id
    if (is.null(dge_base$genes) || !("gene_name" %in% colnames(dge_base$genes))) {
        stop("dge_base$genes must contain a gene_name column.")
    }
    matches <- rownames(dge_base$genes)[dge_base$genes$gene_name == outcome_gene]
    if (length(matches) == 0) stop("Outcome gene not found in dge_base: ", outcome_gene)
    out_gid <- matches[1]
    out_gname <- outcome_gene

    ## get outcome expression (logCPM)
    y_logcpm <- edgeR::cpm(dge_base, log = TRUE)
    y_mat <- matrix(y_logcpm[out_gid, ], nrow = 1)
    rownames(y_mat) <- out_gid
    colnames(y_mat) <- colnames(y_logcpm)

    ## align mediator to outcome samples
    dge_brnum <- gsub("_Oligo\\.3$", "", colnames(dge_base))
    med_vec_aligned <- med_expr[med_gid, dge_brnum]

    ## build design matrices
    baseline_mod <- model.matrix(baseline_formula, data = dge_base$samples)

    sample_df_med <- dge_base$samples
    sample_df_med$med_vec <- med_vec_aligned
    med_mod <- model.matrix(med_formula, data = sample_df_med)

    ## apply cleaningY (preserve APOE, remove covariates)
    y_cleaned_baseline <- jaffelab::cleaningY(y_mat, baseline_mod, P = n_apoe_cols)
    y_cleaned_mediation <- jaffelab::cleaningY(y_mat, med_mod, P = n_apoe_cols)

    ## create plot data frame
    apoe_status <- dge_base$samples$APOE_syn
    carrier_group <- ifelse(apoe_status %in% c("E2.E2", "E2.E3"), "E2+",
                        ifelse(apoe_status %in% c("E3.E4", "E4.E4"), "E4+", "other"))

    plot_df <- data.frame(
        BrNum = dge_brnum,
        carrier_group = carrier_group,
        baseline_cleaned = as.numeric(y_cleaned_baseline[1, ]),
        med_cleaned = as.numeric(y_cleaned_mediation[1, ])
    )

    plot_df <- plot_df[plot_df$carrier_group %in% c("E2+", "E4+"), ]

    model_labels <- c(
        sprintf("Baseline\n(no %s)", mediator_gene),
        sprintf("After %s adjustment", mediator_gene)
    )

    plot_long <- plot_df |>
        tidyr::pivot_longer(
            cols = c(baseline_cleaned, med_cleaned),
            names_to = "model",
            values_to = "expression"
        ) |>
        dplyr::mutate(model = factor(model,
             levels = c("baseline_cleaned", "med_cleaned"),
             labels = model_labels)
        )
    annotation_df <- NULL
    if (!is.null(baseline_results_dt) && !is.null(med_res_dt)) {
        base_row <- baseline_results_dt[outcome_gene_id == out_gid, , drop = FALSE]
        med_row <- med_res_dt[outcome_gene_id == out_gid, , drop = FALSE]
        base_row <- if (nrow(base_row) > 0) base_row[1, ] else NULL
        med_row <- if (nrow(med_row) > 0) med_row[1, ] else NULL

        fdr_base <- if (!is.null(base_row) && "adj.P.Val" %in% names(base_row)) base_row$adj.P.Val else NA_real_
        tvar_base <- if (!is.null(base_row) && "t" %in% names(base_row)) base_row$t else NA_real_
        fdr_med <- if (!is.null(med_row) && "adj.P.Val" %in% names(med_row)) med_row$adj.P.Val else NA_real_
        tvar_med <- if (!is.null(med_row) && "t" %in% names(med_row)) med_row$t else NA_real_

        y_max <- max(plot_long$expression, na.rm = TRUE)
        y_min <- min(plot_long$expression, na.rm = TRUE)
        y_fixed <- y_max + pmax(0.08 * (y_max - y_min), 0.2)

        # Format FDR values: use fixed decimal notation for values >= 1e-4
        format_fdr <- function(fdr) {
            if (is.na(fdr)) return("NA")
            if (fdr >= 1e-4) {
                sprintf("%.5f", fdr)
            } else {
                sprintf("%.2e", fdr)
            }
        }
        annotation_df <- dplyr::tibble(
            model = factor(model_labels, levels = model_labels),
            label = c(
                sprintf("FDR: %s\nt: %.2f", format_fdr(fdr_base), tvar_base),
                sprintf("FDR: %s\nt: %.2f", format_fdr(fdr_med), tvar_med)
            ),
            y = y_fixed
        )
    }
    subtitle <- fifelse(med_region == "ERC",
        sprintf("Mediator: %s expression from %s %s cluster", mediator_gene, med_region, med_cluster),
        ## LC:
        sprintf("Mediator: %s expression from %s spatial domain", mediator_gene, med_cluster)
    )
    ggplot(plot_long, aes(x = carrier_group, y = expression, fill = carrier_group,
                        color = carrier_group)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA) +
        geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
        facet_wrap(~model) +
        scale_fill_manual(values = APOE_carrier_colors) +
        scale_color_manual(values = APOE_carrier_colors_dark) +
        labs( title = sprintf("%s (%s) Expression: E2+ vs E4+ Carriers",
                            out_gname, out_gid),
            subtitle = subtitle,
            x = "APOE Carrier Status",
            y = "Covariate-adjusted Expression (logCPM)",
            fill = "Carrier"
        ) +
        {
            if (!is.null(annotation_df)) {
                ggplot2::geom_text(
                    data = annotation_df,
                    ggplot2::aes(x = 1.5, y = y, label = label),
                    inherit.aes = FALSE,
                    size = 3.5
                )
            }
        } +
        theme_bw(base_size = 14) +
        theme( legend.position = "none",
            strip.text = element_text(size = 12, face = "bold")
        ) # +  coord_cartesian(clip = "off")
}

#p <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", med_region = "LC")
p <- plotMedComp(outcome_gene = "MBP", mediator_gene = "ABCA8")
print(p)
```

### Comparison boxplots BEFORE vs AFTER mediation

LEFT panel (Baseline): E2 vs E4 boxplots showing Oligo.3 gene expression before adding the mediator (Astro PLP1) as a covariate. The expression values are "cleaned" using jaffelab::cleaningY to remove effects of nuisance covariates (Sex, Age, Anc_Afr, pseudo_expr_chrM_ratio) while preserving the APOE effect.

RIGHT panel (After mediation): E2 vs E4 boxplots showing the same outcome gene expression, but now cleaned to remove both nuisance covariates AND the mediator gene expression. If the mediator explains part of the APOE→Oligo.3 effect, the E2 vs E4 difference should shrink.

How cleaningY(y, mod, P) works:
Fits y = mod %*% beta (full model)
Returns y - (effects from columns P+1 to end)
With P=4, preserves APOE columns (1-4), removes covariate effects
Input data needed:
 Outcome expression matrix (y_logcpm): logCPM from edgeR::cpm(dge_base, log=TRUE) for Oligo.3
 Mediator expression vector: From med_expr[med_gene_id, ] aligned to outcome samples
 Design matrices:
  Baseline: ~0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio
  Mediation: same + med_vec
  Sample metadata: dge_base$samples$APOE_syn to group E2.E2 vs E3.E4/E4.E4


```{r LC mediation data load and plot}
## loading LC mediation data
 loadMedData(file.path(cdir, "22_Mediation", "out-lc_astro"))
 plc <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", med_region = "LC")
print(plc)
```








```{r plots in PDFs for specific genes - ERC Astro case}
dnr=c('PLP1', 'MAG', 'MAL', 'MBP', 'SOX10', 'OPALIN')
upr=c('FOS', 'TLR2', 'STAT1', 'STAT4')
alldegs=c(dnr, upr)
medgenes=c('PLP1', 'MBP', 'OPALIN')
pdf_dir <- out_dt
pdf_files <- file.path(pdf_dir, sprintf("mediator_%s.pdf", medgenes))

plot_one_mediator <- function(mediator_gene, outcome_genes, out_file) {
    grDevices::pdf(out_file, width = 11, height = 8.5, onefile = TRUE)
    on.exit(grDevices::dev.off(), add = TRUE)

    for (outcome_gene in outcome_genes) {
        p <- plotMedComp(outcome_gene = outcome_gene, mediator_gene = mediator_gene)
        print(p)
    }

    invisible(out_file)
}

pdf_out <- mapply(
    plot_one_mediator,
    mediator_gene = medgenes,
    out_file = pdf_files,
    MoreArgs = list(outcome_genes = alldegs),
    SIMPLIFY = FALSE
)

pdf_out

```

```{r load and plot LC mediation examples}



```
