---
title: "Plot mediation screening results"
## format: html
---

```{r}
library(edgeR)
library(limma)
library(SingleCellExperiment)
library(data.table)
library(ggplot2)
if (!requireNamespace("ggh4x", quietly = TRUE)) {
    stop("Package `ggh4x` is required for panel-specific facet strip colors. Install with: install.packages('ggh4x')")
}
library(ggh4x)
library(dplyr)
library(here)
ddir <- here("processed-data")
```



```{r setup}
library(qs2)
library(ggplot2)
library(tidyr)
## assumes 01_mediation_screen.R was already run for the scenarios here, to use the output directories
cdir <- here("code")
`%||%` <- function(x, y) if (is.null(x) || length(x) < 1 || all(is.na(x))) y else x

md5_string <- function(x) {
    tf <- tempfile()
    on.exit(unlink(tf), add = TRUE)
    writeLines(x, con = tf, useBytes = TRUE)
    as.character(tools::md5sum(tf)[1])
}

validate_med_expr_src <- function(med_expr_src) {
    if (is.null(med_expr_src)) return(NULL)
    if (!is.list(med_expr_src)) {
        stop("`med_expr_src` must be a list with at least `kind` and `path`.")
    }
    kind <- as.character(med_expr_src$kind %||% "")
    path <- as.character(med_expr_src$path %||% "")
    meta <- med_expr_src$meta %||% list()
    if (!kind %in% c("erc_sce", "lc_sp_astro")) {
        stop("`med_expr_src$kind` must be one of: erc_sce, lc_sp_astro")
    }
    if (!nzchar(path)) {
        stop("`med_expr_src$path` must be a non-empty path.")
    }
    if (!file.exists(path)) {
        stop("`med_expr_src$path` does not exist: ", path)
    }
    if (!is.list(meta)) {
        stop("`med_expr_src$meta` must be a list when provided.")
    }
    list(
        kind = kind,
        path = normalizePath(path, winslash = "/", mustWork = TRUE),
        meta = meta
    )
}

read_med_expr_src <- function(med_expr_src) {
    if (is.null(med_expr_src)) return(NULL)
    src_path <- as.character(med_expr_src$path)
    ## auto-detect serialized object format by file extension
    if (grepl("\\.qs2$", src_path, ignore.case = TRUE)) {
        return(qs_read(src_path))
    }
    readRDS(src_path)
}

file_fingerprint <- function(path) {
    info <- file.info(path)
    if (nrow(info) < 1 || is.na(info$size[1]) || is.na(info$mtime[1])) {
        return(sprintf("%s|missing", normalizePath(path, winslash = "/", mustWork = FALSE)))
    }
    sprintf(
        "%s|%s|%s",
        normalizePath(path, winslash = "/", mustWork = FALSE),
        as.character(info$size[1]),
        format(info$mtime[1], tz = "UTC", usetz = TRUE)
    )
}

loadMedData <- function(out_dir, med_expr_src = NULL) {
    pth <- list(
        out = out_dir,
        base_dir = file.path(out_dir, "baseline"),
        fn_md = file.path(out_dir, "med_degs.qs2"),
        fn_me = file.path(out_dir, "med_expr.qs2"),
        fn_dge = file.path(out_dir, "dge_base.qs2"),
        fn_mi = file.path(out_dir, "mediation_run_index.tsv"),
        fn_bi = file.path(out_dir, "baseline_index.tsv")
    )
    if (!file.exists(pth$fn_md)) stop("Missing `med_degs.qs2`: ", pth$fn_md)
    if (!file.exists(pth$fn_me)) stop("Missing `med_expr.qs2`: ", pth$fn_me)
    if (!file.exists(pth$fn_dge)) stop("Missing `dge_base.qs2`: ", pth$fn_dge)

    md <- data.table::as.data.table(qs_read(pth$fn_md))
    if (!is.data.frame(md) || nrow(md) < 1) {
        stop("`med_degs` is empty or invalid in: ", pth$fn_md)
    }
    me <- qs_read(pth$fn_me)
    if (!is.matrix(me) || nrow(me) < 1) {
        stop("`med_expr` is empty or invalid in: ", pth$fn_me)
    }
    dge <- qs_read(pth$fn_dge)

    mi <- if (file.exists(pth$fn_mi)) data.table::fread(pth$fn_mi) else NULL
    bi <- if (file.exists(pth$fn_bi)) data.table::fread(pth$fn_bi) else NULL

    bf <- NULL
    if (!is.null(bi) && nrow(bi) > 0 && "out_file" %in% names(bi)) {
        bi2 <- data.table::copy(bi)
        if ("n_donors" %in% names(bi2)) {
            data.table::setorderv(bi2, cols = "n_donors", order = -1L)
        }
        bf <- bi2$out_file[1]
    } else {
        bf <- list.files(pth$base_dir, pattern = "^baseline_[0-9a-f]+\\.tab\\.gz$", full.names = TRUE)[1]
    }
    if (is.na(bf) || !nzchar(bf) || !file.exists(bf)) {
        stop("Could not resolve baseline file in: ", pth$base_dir)
    }
    br <- data.table::fread(bf)
    if (!is.data.frame(br) || nrow(br) < 1) {
        stop("`baseline_results_dt` is empty in: ", bf)
    }

    if (is.null(dge$genes) || !("gene_name" %in% colnames(dge$genes))) {
        stop("`dge_base$genes` must contain `gene_name`.")
    }
    gmap <- data.table::data.table(
        gene_name = as.character(dge$genes$gene_name),
        gene_id = rownames(dge$genes)
    )
    brnum <- gsub("_Oligo\\.3$", "", colnames(dge))
    ap <- as.character(dge$samples$APOE_syn)
    grp <- ifelse(ap %in% c("E2.E2", "E2.E3"), "E2+",
        ifelse(ap %in% c("E3.E4", "E4.E4"), "E4+", "other")
    )

    cfg <- list(
        base_str = "0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio",
        n_apoe = 4L
    )
    cfg$med_str <- paste(cfg$base_str, "+ med_vec")
    cfg$base_f <- as.formula(paste("~", cfg$base_str))
    cfg$med_f <- as.formula(paste("~", cfg$med_str))

    med_expr_src <- validate_med_expr_src(med_expr_src)
    src <- list(
        med_expr_src = med_expr_src,
        src_obj = read_med_expr_src(med_expr_src)
    )

    cfg <- c(cfg, list(
        apoe_levels = c("E2.E2", "E2.E3", "E3.E4", "E4.E4"),
        carrier_contrast = "-0.5*(APOE_E2.E2 + APOE_E2.E3) + 0.5*(APOE_E3.E4 + APOE_E4.E4)",
        erc_batch_col = "exp_round"
    ))

    pd <- list(
        pth = pth,
        dt = list(md = md, mi = mi, bi = bi, br = br),
        obj = list(dge = dge),
        mat = list(y_lcpm = edgeR::cpm(dge, log = TRUE), m_expr = me),
        key = list(gmap = gmap, brnum = brnum, grp = grp),
        cfg = cfg,
        src = src
    )
    pd
}


```

```{r load mediation data}
## loading ERC Astro mediation data
erc_src_sce <- file.path(ddir, "08_pseudoBulkDGE_sn", "01_pseudobulk_data_sn", "sce_pseudo_DGE-cell_type_anno.RDS")
pd <- loadMedData(
    file.path(cdir, "22_Mediation", "out-erc_astro"),
    med_expr_src = list(kind = "erc_sce", path = erc_src_sce)
)
```


```{r plot function and test}
APOE_carrier_colors_dark <- c(`E2+`="#3B918B", `E4+`="#B7522A")
APOE_carrier_colors <- c(`E2+`="#51B8B1", `E4+`="#D97D59")

carrier_group_from_apoe <- function(apoe_syn) {
    ap <- as.character(apoe_syn)
    ifelse(ap %in% c("E2.E2", "E2.E3"), "E2+",
        ifelse(ap %in% c("E3.E4", "E4.E4"), "E4+", "other")
    )
}

format_fdr <- function(fdr) {
    if (is.na(fdr)) return("NA")
    if (fdr >= 1e-4) sprintf("%.5f", fdr) else sprintf("%.2e", fdr)
}

format_tstat <- function(tval) {
    if (is.na(tval)) return("NA")
    sprintf("%.2f", tval)
}

extract_gene_stats <- function(dt, gene_id, gene_col = "outcome_gene_id") {
    if (is.null(dt) || !is.data.frame(dt) || nrow(dt) < 1 || !(gene_col %in% names(dt))) {
        return(list(fdr = NA_real_, t = NA_real_))
    }
    hit_idx <- which(as.character(dt[[gene_col]]) == gene_id)
    if (length(hit_idx) < 1) {
        return(list(fdr = NA_real_, t = NA_real_))
    }
    row <- dt[hit_idx[1], ]
    list(
        fdr = if ("adj.P.Val" %in% names(row)) as.numeric(row$adj.P.Val[[1]]) else NA_real_,
        t = if ("t" %in% names(row)) as.numeric(row$t[[1]]) else NA_real_
    )
}

resolve_mediator_row <- function(mg, md) {
    mg_cl <- NULL
    mg_nm <- mg
    if (grepl(":", mg, fixed = TRUE)) {
        pr <- strsplit(mg, ":", fixed = TRUE)[[1]]
        mg_cl <- pr[1]
        mg_nm <- pr[2]
    }
    rs <- md[
        gene_name == mg_nm | gene_id == mg_nm | med_gene_id == mg_nm
    ]
    if (!is.null(mg_cl)) {
        rs <- rs[cluster == mg_cl]
    }
    if (nrow(rs) == 0) {
        cands <- unique(md$gene_name)
        cls <- cands[agrep(mg_nm, cands, max.distance = 0.2)]
        stop("Mediator not found: ", mg, "\n",
            "Tried gene_name / gene_id / med_gene_id.\n",
            "Close matches: ", paste(head(cls, 6), collapse = ", "))
    }
    if (nrow(rs) > 1) {
        warning(sprintf(
            "Multiple matches for mediator gene '%s' in clusters: %s. Using last match; use 'cluster:gene' to disambiguate.",
            mg, paste(unique(rs$cluster), collapse = ", ")
        ))
    }
    rs[nrow(rs)]
}

resolve_mediator_results_file <- function(pd, dt_mi, it, mgid, file_label) {
    mf <- character(0)
    if (!is.null(dt_mi) && nrow(dt_mi) > 0 &&
        all(c("iter", "med_gene_id", "out_file") %in% names(dt_mi))) {
        idx <- dt_mi[iter == it & med_gene_id == mgid]
        if (nrow(idx) > 0) mf <- idx$out_file[1]
    }
    if (length(mf) < 1 || !file.exists(mf)) {
        mf <- file.path(pd$pth$out, sprintf("%03d_%s.tab.gz", it, file_label))
    }
    if (!file.exists(mf)) {
        cands <- list.files(pd$pth$out, pattern = sprintf("_%s\\.tab\\.gz$", file_label), full.names = TRUE)
        if (length(cands) > 1) {
            finfo <- file.info(cands)
            cands <- cands[order(finfo$mtime, decreasing = TRUE)]
        }
        mf <- cands[1]
    }
    if (is.na(mf) || !nzchar(mf) || !file.exists(mf)) {
        stop("Mediation results file not found for mediator run label: ", file_label)
    }
    mf
}

get_precomputed_med_stats <- function(mr) {
    fdr_cols <- c("vlmf_adj.P.Val", "adj.P.Val")
    t_cols <- c("vlmf_t", "t")
    fdr_col <- fdr_cols[fdr_cols %in% names(mr)][1]
    t_col <- t_cols[t_cols %in% names(mr)][1]
    list(
        fdr = if (!is.na(fdr_col)) as.numeric(mr[[fdr_col]][1]) else NA_real_,
        t = if (!is.na(t_col)) as.numeric(mr[[t_col]][1]) else NA_real_
    )
}

get_source_object_for_kind <- function(pd, required_kind) {
    src_cfg <- pd$src$med_expr_src %||% NULL
    if (is.null(src_cfg)) {
        stop("No `med_expr_src` configured in plot data. Call loadMedData(..., med_expr_src=list(kind=..., path=...)).")
    }
    if (as.character(src_cfg$kind) != as.character(required_kind)) {
        stop(sprintf("`med_expr_src$kind` is '%s' but '%s' is required for this operation.", src_cfg$kind, required_kind))
    }
    src_obj <- pd$src$src_obj %||% NULL
    if (!is.null(src_obj)) return(src_obj)
    if (!file.exists(src_cfg$path)) {
        stop("Source object path not found: ", src_cfg$path)
    }
    read_med_expr_src(src_cfg)
}

build_med_source_context_erc <- function(mr, pd, brnum_use, med_vec_use, n_ap) {
    sce_src <- get_source_object_for_kind(pd, "erc_sce")
    mcl <- as.character(mr$cluster)
    cl_idx <- as.character(SummarizedExperiment::colData(sce_src)$registration_variable) == mcl
    if (!any(cl_idx)) {
        stop("Cluster not found in source ERC SCE: ", mcl)
    }
    cl_sce <- sce_src[, cl_idx]
    cl_smp <- as.data.frame(SummarizedExperiment::colData(cl_sce))
    cl_smp$BrNum <- as.character(cl_smp$BrNum)
    if (anyDuplicated(cl_smp$BrNum)) {
        stop("Duplicate BrNum entries found in source ERC SCE for cluster: ", mcl)
    }
    ord <- match(brnum_use, cl_smp$BrNum)
    if (anyNA(ord)) {
        miss <- brnum_use[is.na(ord)]
        stop(sprintf("Missing cluster samples for %s, first missing donor(s): %s",
                     mcl, paste(head(miss, 5), collapse = ", ")))
    }
    cl_smp <- cl_smp[ord, , drop = FALSE]
    cl_smp$APOE_syn <- factor(as.character(cl_smp$APOE_syn), levels = pd$cfg$apoe_levels)
    cl_smp$Sex <- droplevels(factor(cl_smp$Sex))
    mod_step2 <- model.matrix(pd$cfg$base_f, data = cl_smp)

    m_mat <- matrix(med_vec_use, nrow = 1, dimnames = list(as.character(mr$gene_id), brnum_use))
    m_clean <- jaffelab::cleaningY(m_mat, mod_step2, P = n_ap)
    data.frame(
        BrNum = brnum_use,
        carrier_group = carrier_group_from_apoe(cl_smp$APOE_syn),
        med_cleaned = as.numeric(m_clean[1, ])
    )
}

build_med_source_context_outcome <- function(mr, brnum_use, med_vec_use, smp, n_ap, mod_b) {
    m_mat <- matrix(med_vec_use, nrow = 1, dimnames = list(as.character(mr$gene_id), brnum_use))
    m_clean <- jaffelab::cleaningY(m_mat, mod_b, P = n_ap)
    data.frame(
        BrNum = brnum_use,
        carrier_group = carrier_group_from_apoe(smp$APOE_syn),
        med_cleaned = as.numeric(m_clean[1, ])
    )
}

resolve_lc_donor_pb_path <- function(pd) {
    src_cfg <- pd$src$med_expr_src %||% NULL
    if (is.null(src_cfg) || as.character(src_cfg$kind) != "lc_sp_astro") {
        stop("LC donor pseudobulk path resolution requires `med_expr_src$kind='lc_sp_astro'`.")
    }
    meta <- src_cfg$meta %||% list()
    cand <- character(0)
    if (!is.null(meta$donor_pb_path) && nzchar(as.character(meta$donor_pb_path))) {
        cand <- c(cand, as.character(meta$donor_pb_path))
    }
    ## try known default donor-pseudobulk source paths before failing
    cand <- c(
        cand,
        file.path(here("processed-data"), "22_Mediation", "lc_astro_pseudobulk_counts_by_donor.rds"),
        file.path(dirname(src_cfg$path), "shared_internal_ref", "DGEList_pseudobulkcounts_by_donor_forGP_unfilt_020426.RDS"),
        file.path(here("code"), "22_Mediation", "LC-processed-data", "shared_internal_ref", "DGEList_pseudobulkcounts_by_donor_forGP_unfilt_020426.RDS")
    )
    cand <- unique(cand[nzchar(cand)])
    cand_hit <- cand[file.exists(cand)]
    if (length(cand_hit) < 1L) {
        stop(
            "Could not find LC donor-pseudobulk counts for recalc. ",
            "Provide `med_expr_src$meta$donor_pb_path` (e.g., .../lc_astro_pseudobulk_counts_by_donor.rds ",
            "or .../shared_internal_ref/DGEList_pseudobulkcounts_by_donor_forGP_unfilt_020426.RDS)."
        )
    }
    normalizePath(cand_hit[1], winslash = "/", mustWork = TRUE)
}

read_by_ext <- function(path) {
    ## support both rds and qs2 serialized files
    if (grepl("\\.qs2$", path, ignore.case = TRUE)) {
        return(qs_read(path))
    }
    readRDS(path)
}

pick_lc_gene_row <- function(mr, logcpm, genes_df = NULL) {
    gid <- as.character(mr$gene_id)
    gnm <- as.character(mr$gene_name)
    if (gid %in% rownames(logcpm)) return(gid)
    if (!is.null(genes_df) && "gene_id" %in% names(genes_df) && !is.null(rownames(genes_df))) {
        map <- rownames(genes_df)[as.character(genes_df$gene_id) == gid]
        if (length(map) > 0 && map[1] %in% rownames(logcpm)) return(map[1])
    }
    if (!is.null(genes_df) && "gene_name" %in% names(genes_df) && !is.null(rownames(genes_df))) {
        map <- rownames(genes_df)[as.character(genes_df$gene_name) == gnm]
        map <- map[map %in% rownames(logcpm)]
        if (length(map) > 0) return(map[1])
    }
    NA_character_
}

build_med_source_context_lc_precomputed <- function(mr, pd, n_ap = 2L) {
    src_obj <- get_source_object_for_kind(pd, "lc_sp_astro")
    if (is.null(src_obj$logCPM) || is.null(src_obj$targets)) {
        stop("LC source object must contain `logCPM` and `targets` for precomputed panel-3 plotting.")
    }
    logcpm <- src_obj$logCPM
    targets <- as.data.frame(src_obj$targets)
    if (!("label" %in% names(targets))) {
        stop("LC source `targets` is missing `label` column.")
    }

    ## subset to Astro tissue-section samples from the original LC model context
    astro_idx <- as.character(targets$label) == "Astro"
    if (!any(astro_idx)) {
        stop("No Astro section samples found in LC source `targets$label`.")
    }
    astro_targets <- targets[astro_idx, , drop = FALSE]
    astro_cols <- colnames(logcpm)[astro_idx]
    astro_targets$label_E2E4 <- factor(as.character(astro_targets$label_E2E4))
    astro_targets$Sex <- droplevels(factor(astro_targets$Sex))

    ## residualize section-level mediator expression while preserving Astro E2/E4 effects
    mod_step2 <- model.matrix(~0 + label_E2E4 + Sex + Age + YRI, data = astro_targets)
    colnames(mod_step2) <- gsub("^label_E2E4", "", colnames(mod_step2))
    if (!all(c("Astro_E2", "Astro_E4") %in% colnames(mod_step2))) {
        stop("LC Astro section design does not contain expected APOE columns `Astro_E2` and `Astro_E4`.")
    }

    gene_row <- pick_lc_gene_row(mr, logcpm, src_obj$genes %||% NULL)
    if (is.na(gene_row) || !nzchar(gene_row)) {
        stop(sprintf("Mediator %s (%s) not found in LC source logCPM matrix.",
                     as.character(mr$gene_name), as.character(mr$gene_id)))
    }
    med_vec <- as.numeric(logcpm[gene_row, astro_cols])
    m_mat <- matrix(med_vec, nrow = 1, dimnames = list(as.character(mr$gene_id), astro_cols))
    m_clean <- jaffelab::cleaningY(m_mat, mod_step2, P = as.integer(n_ap))

    brnum <- as.character(astro_targets$brnum %||% rep(NA_character_, nrow(astro_targets)))
    brnum[is.na(brnum) | !nzchar(brnum)] <- astro_cols[is.na(brnum) | !nzchar(brnum)]
    ap_large <- as.character(astro_targets$APOlargeGrp %||% NA_character_)
    if (length(ap_large) != nrow(astro_targets) || all(is.na(ap_large))) {
        ap_large <- ifelse(grepl("_E2$", as.character(astro_targets$label_E2E4)), "E2",
                           ifelse(grepl("_E4$", as.character(astro_targets$label_E2E4)), "E4", NA_character_))
    }
    grp <- ifelse(ap_large == "E2", "E2+",
                  ifelse(ap_large == "E4", "E4+", "other"))
    data.frame(
        BrNum = brnum,
        carrier_group = grp,
        med_cleaned = as.numeric(m_clean[1, ])
    )
}

get_med_source_context <- function(mr, pd, brnum_use, med_vec_use, n_ap, smp, mod_b, med_mode_eff) {
    src_cfg <- pd$src$med_expr_src %||% NULL
    if (is.null(src_cfg)) {
        return(build_med_source_context_outcome(mr, brnum_use, med_vec_use, smp, n_ap, mod_b))
    }
    src_kind <- as.character(src_cfg$kind)
    if (src_kind=="erc_sce") {
        return(build_med_source_context_erc(mr, pd, brnum_use, med_vec_use, n_ap))
    }
    if (src_kind == "lc_sp_astro") {
        ## route LC panel-3 points by mode: section-level for precomputed/both, donor-level for recalc
        if (med_mode_eff %in% c("precomputed", "both")) {
            return(build_med_source_context_lc_precomputed(mr, pd, n_ap = 2L))
        }
        return(build_med_source_context_outcome(mr, brnum_use, med_vec_use, smp, n_ap, mod_b))
    }
    stop("Unsupported med_expr source kind: ", src_kind)
}

med_cache_key_erc <- function(pd, cluster) {
    src_cfg <- pd$src$med_expr_src %||% NULL
    if (is.null(src_cfg) || as.character(src_cfg$kind) != "erc_sce") {
        stop("Cannot build ERC Step-2 cache key without `med_expr_src` kind='erc_sce'.")
    }
    info <- file.info(src_cfg$path)
    if (nrow(info) < 1 || is.na(info$size[1]) || is.na(info$mtime[1])) {
        stop("Cannot stat ERC source path for cache key: ", src_cfg$path)
    }
    key_txt <- paste(
        c(
            "erc_med_exact_full_cluster",
            normalizePath(src_cfg$path, winslash = "/", mustWork = FALSE),
            as.character(info$size[1]),
            format(info$mtime[1], tz = "UTC", usetz = TRUE),
            as.character(cluster),
            as.character(pd$cfg$base_str),
            as.character(pd$cfg$carrier_contrast),
            as.character(pd$cfg$erc_batch_col),
            as.character(utils::packageVersion("edgeR")),
            as.character(utils::packageVersion("limma"))
        ),
        collapse = "\n"
    )
    md5_string(key_txt)
}

get_or_compute_erc_cluster_dge <- function(cluster, pd, med_cache_dir) {
    if (!dir.exists(med_cache_dir)) dir.create(med_cache_dir, recursive = TRUE, showWarnings = FALSE)
    cache_file <- file.path(
        med_cache_dir,
        sprintf("erc_med_%s.qs2", gsub("[^A-Za-z0-9._-]", "_", as.character(cluster)))
    )
    cache_key <- med_cache_key_erc(pd, cluster)
    if (file.exists(cache_file)) {
        cached <- tryCatch(qs_read(cache_file), error = function(e) NULL)
        if (is.list(cached) && identical(cached$cache_key %||% "", cache_key) &&
            is.data.frame(cached$table)) {
            return(data.table::as.data.table(cached$table))
        }
    }

    sce_src <- get_source_object_for_kind(pd, "erc_sce")
    cl_idx <- as.character(SummarizedExperiment::colData(sce_src)$registration_variable) == cluster
    if (!any(cl_idx)) {
        stop("Cluster not found in ERC source SCE for recalc: ", cluster)
    }
    dge <- sce_src[, cl_idx]
    smp <- as.data.frame(SummarizedExperiment::colData(dge))
    smp$APOE_syn <- factor(as.character(smp$APOE_syn), levels = pd$cfg$apoe_levels)
    smp$Sex <- droplevels(factor(smp$Sex))

    des <- model.matrix(pd$cfg$base_f, data = smp)
    des <- as.data.frame(des)
    colnames(des) <- gsub(colnames(des), pattern = "_syn", replacement = "_")

    dge <- edgeR::calcNormFactors(dge)
    keep <- edgeR::filterByExpr(dge, design = des)
    dge <- dge[keep, , keep.lib.sizes = FALSE]
    dge <- edgeR::calcNormFactors(dge)
    ## report the exact dataset used when recomputing mediator dge
    message(sprintf(
        "[med-recalc][erc] cluster=%s samples=%d rows=%d model=%s",
        as.character(cluster), ncol(dge), nrow(dge), as.character(pd$cfg$base_f)
    ))

    batch_col <- as.character(pd$cfg$erc_batch_col %||% "exp_round")
    if (!(batch_col %in% colnames(dge$samples))) {
        stop("Batch column not found in ERC source dge$samples: ", batch_col)
    }
    v.swt <- edgeR::voomLmFit(
        dge,
        design = des,
        block = as.factor(dge$samples[[batch_col]]),
        adaptive.span = TRUE,
        sample.weights = TRUE
    )

    cont <- limma::makeContrasts(
        contrasts = pd$cfg$carrier_contrast,
        levels = des
    )
    colnames(cont) <- "carrier"
    fit <- limma::eBayes(limma::contrasts.fit(v.swt, contrasts = cont))
    tt <- limma::topTable(fit, coef = 1, number = Inf, adjust.method = "BH")
    tt$gene_id <- rownames(tt)
    tt <- data.table::as.data.table(tt)
    if (!is.null(dge$genes) && "gene_name" %in% colnames(dge$genes)) {
        gnm <- as.character(dge$genes$gene_name)
        names(gnm) <- rownames(dge$genes)
        tt[, gene_name := unname(gnm[gene_id])]
    }

    qs_save(list(
        cache_key = cache_key,
        cluster = as.character(cluster),
        created_at = format(Sys.time(), tz = "UTC", usetz = TRUE),
        table = tt
    ), cache_file)
    tt
}

get_recalc_med_stats_erc <- function(mr, pd, med_cache_dir) {
    tt <- get_or_compute_erc_cluster_dge(
        cluster = as.character(mr$cluster),
        pd = pd,
        med_cache_dir = med_cache_dir
    )
    hit <- tt[gene_id == as.character(mr$gene_id)]
    if (nrow(hit) < 1 && "gene_name" %in% names(tt)) {
        hit <- tt[gene_name == as.character(mr$gene_name)]
    }
    if (nrow(hit) < 1) {
        warning(sprintf("Step-2 recalc could not find mediator %s (%s) in cluster %s.",
                        as.character(mr$gene_name), as.character(mr$gene_id), as.character(mr$cluster)))
        return(list(fdr = NA_real_, t = NA_real_))
    }
    list(
        fdr = as.numeric(hit$adj.P.Val[[1]]),
        t = as.numeric(hit$t[[1]])
    )
}

med_cache_key_lc <- function(pd, donor_pb_path, brnum_use) {
    ## encode source, donor set, and model settings to invalidate stale LC recalc caches
    key_txt <- paste(
        c(
            "lc_med_donor_recalc",
            file_fingerprint(donor_pb_path),
            paste(sort(unique(as.character(brnum_use))), collapse = ","),
            "~0 + APOE_syn + Sex + Age + YRI",
            as.character(pd$cfg$carrier_contrast),
            as.character(utils::packageVersion("edgeR")),
            as.character(utils::packageVersion("limma"))
        ),
        collapse = "\n"
    )
    md5_string(key_txt)
}

extract_lc_astro_donor_counts <- function(dgl) {
    astro_cols <- grep("^Astro_", colnames(dgl), value = TRUE)
    if (length(astro_cols) > 0) {
        out <- dgl[, astro_cols, keep.lib.sizes = FALSE]
        colnames(out) <- sub("^Astro_", "", astro_cols)
        return(out)
    }
    smp <- dgl$samples
    if (!is.null(smp) && "label" %in% names(smp) && any(as.character(smp$label) == "Astro")) {
        astro_idx <- as.character(smp$label) == "Astro"
        out <- dgl[, astro_idx, keep.lib.sizes = FALSE]
        if ("brnum" %in% names(out$samples) && !all(is.na(out$samples$brnum))) {
            colnames(out) <- as.character(out$samples$brnum)
        }
        return(out)
    }
    stop("Could not resolve Astro donor columns in LC donor-pseudobulk DGEList.")
}

build_lc_recalc_table <- function(pd, donor_pb_path, brnum_use) {
    ## load donor-pseudobulk LC counts and retain Astro domain donors used in this mediator plot
    dgl <- read_by_ext(donor_pb_path)
    if (!inherits(dgl, "DGEList")) {
        stop("LC donor-pseudobulk source must load to an edgeR `DGEList`: ", donor_pb_path)
    }
    dge <- extract_lc_astro_donor_counts(dgl)

    if (anyDuplicated(colnames(dge))) {
        dups <- unique(colnames(dge)[duplicated(colnames(dge))])
        stop("Duplicate donor columns detected in LC donor-pseudobulk counts: ",
             paste(head(dups, 5), collapse = ", "))
    }
    miss <- setdiff(brnum_use, colnames(dge))
    if (length(miss) > 0) {
        stop("Missing donor(s) in LC donor-pseudobulk counts, first missing: ",
             paste(head(miss, 5), collapse = ", "))
    }
    dge <- dge[, brnum_use, keep.lib.sizes = FALSE]

    ## build donor metadata from outcome dge sample table, mapping Anc_Afr to YRI where needed
    smp_ref <- as.data.frame(pd$obj$dge$samples)
    smp_ref$BrNum <- as.character(pd$key$brnum)
    ord <- match(brnum_use, smp_ref$BrNum)
    if (anyNA(ord)) {
        stop("Could not align one or more LC donor(s) to outcome sample metadata for recalc.")
    }
    smp <- smp_ref[ord, , drop = FALSE]
    smp$APOE_syn <- factor(as.character(smp$APOE_syn), levels = pd$cfg$apoe_levels)
    smp$Sex <- droplevels(factor(smp$Sex))
    smp$YRI <- as.numeric(smp$YRI %||% smp$Anc_Afr)
    if (anyNA(smp$YRI)) {
        stop("Missing ancestry values when constructing LC recalc design (`YRI`).")
    }
    des <- model.matrix(~0 + APOE_syn + Sex + Age + YRI, data = smp)
    des <- as.data.frame(des)
    colnames(des) <- gsub(colnames(des), pattern = "_syn", replacement = "_")

    ## run donor-level LC recalc DGE under the carrier contrast used in mediation models
    dge <- edgeR::calcNormFactors(dge)
    keep <- edgeR::filterByExpr(dge, design = des)
    dge <- dge[keep, , keep.lib.sizes = FALSE]
    dge <- edgeR::calcNormFactors(dge)
    ## report the exact donor-level dataset used when recomputing lc mediator dge
    message(sprintf(
        "[med-recalc][lc] donors=%d rows=%d model=%s",
        ncol(dge), nrow(dge), "~0 + APOE_syn + Sex + Age + YRI"
    ))
    v.swt <- edgeR::voomLmFit(dge, design = des, adaptive.span = TRUE, sample.weights = TRUE)
    cont <- limma::makeContrasts(contrasts = pd$cfg$carrier_contrast, levels = des)
    colnames(cont) <- "carrier"
    fit <- limma::eBayes(limma::contrasts.fit(v.swt, contrasts = cont))
    tt <- limma::topTable(fit, coef = 1, number = Inf, adjust.method = "BH")
    tt$gene_id <- rownames(tt)
    tt <- data.table::as.data.table(tt)
    if (!is.null(dge$genes) && "gene_name" %in% colnames(dge$genes)) {
        gnm <- as.character(dge$genes$gene_name)
        names(gnm) <- rownames(dge$genes)
        tt[, gene_name := unname(gnm[gene_id])]
    }
    tt
}

get_or_compute_lc_donor_dge <- function(pd, brnum_use, med_cache_dir) {
    if (!dir.exists(med_cache_dir)) dir.create(med_cache_dir, recursive = TRUE, showWarnings = FALSE)

    donor_pb_path <- resolve_lc_donor_pb_path(pd)
    donor_key <- md5_string(paste(sort(unique(as.character(brnum_use))), collapse = ","))
    cache_file <- file.path(med_cache_dir, sprintf("lc_med_%s.qs2", donor_key))
    cache_key <- med_cache_key_lc(pd, donor_pb_path, brnum_use)
    if (file.exists(cache_file)) {
        cached <- tryCatch(qs_read(cache_file), error = function(e) NULL)
        if (is.list(cached) && identical(cached$cache_key %||% "", cache_key) &&
            is.data.frame(cached$table)) {
            return(data.table::as.data.table(cached$table))
        }
    }

    tt <- build_lc_recalc_table(pd, donor_pb_path, brnum_use)
    qs_save(list(
        cache_key = cache_key,
        donor_pb_path = donor_pb_path,
        donors = as.character(brnum_use),
        created_at = format(Sys.time(), tz = "UTC", usetz = TRUE),
        table = tt
    ), cache_file)
    tt
}

get_recalc_med_stats_lc <- function(mr, pd, brnum_use, med_cache_dir) {
    tt <- get_or_compute_lc_donor_dge(
        pd = pd,
        brnum_use = brnum_use,
        med_cache_dir = med_cache_dir
    )
    hit <- tt[gene_id == as.character(mr$gene_id)]
    if (nrow(hit) < 1 && "gene_name" %in% names(tt)) {
        hit <- tt[gene_name == as.character(mr$gene_name)]
    }
    if (nrow(hit) < 1) {
        warning(sprintf("LC recalc could not find mediator %s (%s).",
                        as.character(mr$gene_name), as.character(mr$gene_id)))
        return(list(fdr = NA_real_, t = NA_real_))
    }
    list(
        fdr = as.numeric(hit$adj.P.Val[[1]]),
        t = as.numeric(hit$t[[1]])
    )
}

get_recalc_med_stats <- function(mr, pd, med_cache_dir, brnum_use = NULL) {
    src_cfg <- pd$src$med_expr_src %||% NULL
    src_kind <- if (is.null(src_cfg)) "" else as.character(src_cfg$kind)
    if (identical(src_kind, "erc_sce")) {
        return(get_recalc_med_stats_erc(mr, pd, med_cache_dir))
    }
    if (identical(src_kind, "lc_sp_astro")) {
        if (is.null(brnum_use) || length(brnum_use) < 2L) {
            stop("LC recalc requires the mediator donor set (`brnum_use`) from the current plot.")
        }
        return(get_recalc_med_stats_lc(mr, pd, brnum_use, med_cache_dir))
    }
    stop("Step-2 recalc currently requires `med_expr_src=list(kind=\"erc_sce\", path=...)`.")
}

build_med_label <- function(mode_eff, pre_stats, rec_stats) {
    if (mode_eff == "precomputed") {
        return(sprintf("FDR: %s\nt: %s", format_fdr(pre_stats$fdr), format_tstat(pre_stats$t)))
    }
    if (mode_eff == "recalc") {
        return(sprintf("Recalc FDR: %s\nRecalc t: %s", format_fdr(rec_stats$fdr), format_tstat(rec_stats$t)))
    }
    paste(
        sprintf("Recalc FDR: %s", format_fdr(rec_stats$fdr)),
        sprintf("Recalc t: %s", format_tstat(rec_stats$t)),
        sprintf("Precalc FDR: %s", format_fdr(pre_stats$fdr)),
        sprintf("Precalc t: %s", format_tstat(pre_stats$t)),
        sep = "\n"
    )
}
library(ggplot2)

plotMedComp <- function(outcome_gene, mediator_gene, pd, med_region = "ERC", n_apoe_cols = 4L,
                        med_stats_mode = c("precomputed", "recalc", "both"),
                        med_cache_dir = NULL, rows = 1L) {
    if (is.null(pd) || !is.list(pd)) stop("`pd` must be a plotData list from loadMedData().")
    med_stats_mode <- match.arg(med_stats_mode)
    if (length(rows) != 1L || is.na(rows) || !is.numeric(rows) ||
        (rows %% 1 != 0) || !(as.integer(rows) %in% c(1L, 2L))) {
        stop("`rows` must be a scalar integer in {1, 2}.")
    }
    rows <- as.integer(rows)
    if (is.null(med_cache_dir) || !nzchar(med_cache_dir)) {
        med_cache_dir <- file.path(pd$pth$out, "plot_cache")
    }

    dge <- pd$obj$dge
    dt_md <- pd$dt$md
    dt_mi <- pd$dt$mi
    dt_br <- pd$dt$br
    m_expr <- pd$mat$m_expr
    y_lcpm <- pd$mat$y_lcpm
    gmap <- pd$key$gmap
    brnum <- pd$key$brnum
    grp <- pd$key$grp
    base_f <- pd$cfg$base_f
    med_f <- pd$cfg$med_f
    n_ap <- as.integer(n_apoe_cols)

    mr <- resolve_mediator_row(mediator_gene, dt_md)
    run_lb <- sprintf("%s|%s|%s", mr$cluster, mr$gene_id, mr$gene_name)
    f_lb <- gsub("\\|", "_", run_lb)
    it <- as.integer(mr$iter)
    mgid <- as.character(mr$med_gene_id)
    mcl <- as.character(mr$cluster)
    med_mode_eff <- med_stats_mode

    mf <- resolve_mediator_results_file(pd, dt_mi, it, mgid, f_lb)
    dt_mr <- data.table::fread(mf)

    if (!mgid %in% rownames(m_expr)) {
        stop("Mediator expression not found in `pd$mat$m_expr` for: ", mgid)
    }

    og <- outcome_gene
    ghit <- gmap[gene_name == og]
    if (nrow(ghit) < 1) {
        stop("Outcome gene not found in dge_base: ", outcome_gene)
    }
    out_gid <- ghit$gene_id[1]

    med_all <- as.numeric(m_expr[mgid, brnum])
    keep_idx <- which(!is.na(med_all))
    if (length(keep_idx) < 2L) {
        stop("Too few non-NA mediator values available for plotting: ", mgid)
    }
    brnum_use <- brnum[keep_idx]
    grp_use <- grp[keep_idx]
    med_vec <- med_all[keep_idx]

    y_mat <- matrix(y_lcpm[out_gid, keep_idx], nrow = 1)
    rownames(y_mat) <- out_gid
    colnames(y_mat) <- brnum_use

    smp <- as.data.frame(dge$samples[keep_idx, , drop = FALSE])
    smp$APOE_syn <- factor(as.character(smp$APOE_syn), levels = pd$cfg$apoe_levels)
    smp$Sex <- droplevels(factor(smp$Sex))
    mod_b <- model.matrix(base_f, data = smp)
    smp_m <- smp
    smp_m$med_vec <- med_vec
    mod_m <- model.matrix(med_f, data = smp_m)

    yb <- jaffelab::cleaningY(y_mat, mod_b, P = n_ap)
    ym <- jaffelab::cleaningY(y_mat, mod_m, P = n_ap)

    plot_df <- data.frame(
        BrNum = brnum_use,
        carrier_group = grp_use,
        baseline_cleaned = as.numeric(yb[1, ]),
        med_cleaned = as.numeric(ym[1, ])
    )

    if (med_mode_eff == "both") {
        mdl_labs <- c(
            "Outcome (baseline)",
            "Outcome (adjusted)",
            "Mediator (original)",
            "Mediator (re-tested)"
        )
    } else {
        mdl_labs <- c(
            "Outcome (baseline)",
            "Outcome (adjusted)",
            "Mediator (source context)"
        )
    }
    plot_long_xy <- plot_df |>
        tidyr::pivot_longer(
            cols = c(baseline_cleaned, med_cleaned),
            names_to = "model",
            values_to = "expression"
        ) |>
        dplyr::mutate(model = factor(model,
             levels = c("baseline_cleaned", "med_cleaned"),
             labels = mdl_labs[1:2])
        )
    if (med_mode_eff == "both") {
        med_df_orig <- get_med_source_context(
            mr = mr, pd = pd, brnum_use = brnum_use, med_vec_use = med_vec,
            n_ap = n_ap, smp = smp, mod_b = mod_b, med_mode_eff = "precomputed"
        )
        med_df_retest <- build_med_source_context_outcome(
            mr = mr, brnum_use = brnum_use, med_vec_use = med_vec, smp = smp,
            n_ap = n_ap, mod_b = mod_b
        )
        plot_long_med_orig <- med_df_orig |>
            dplyr::transmute(BrNum = BrNum, carrier_group = carrier_group,
                             model = mdl_labs[3], expression = med_cleaned)
        plot_long_med_retest <- med_df_retest |>
            dplyr::transmute(BrNum = BrNum, carrier_group = carrier_group,
                             model = mdl_labs[4], expression = med_cleaned)
        panel3_n <- plot_long_med_orig |>
            dplyr::filter(carrier_group %in% c("E2+", "E4+")) |>
            nrow()
        panel4_n <- plot_long_med_retest |>
            dplyr::filter(carrier_group %in% c("E2+", "E4+")) |>
            nrow()
        message(sprintf(
            "[plotMedComp] mediator panel samples orig=%d retest=%d mode=%s cluster=%s",
            panel3_n, panel4_n, med_mode_eff, as.character(mr$cluster)
        ))
        plot_long <- dplyr::bind_rows(plot_long_xy, plot_long_med_orig, plot_long_med_retest) |>
            dplyr::filter(carrier_group %in% c("E2+", "E4+")) |>
            dplyr::mutate(model = factor(as.character(model), levels = mdl_labs))
    } else {
        med_df <- get_med_source_context(
            mr = mr, pd = pd, brnum_use = brnum_use, med_vec_use = med_vec,
            n_ap = n_ap, smp = smp, mod_b = mod_b, med_mode_eff = med_mode_eff
        )
        plot_long_step2 <- med_df |>
            dplyr::transmute(BrNum = BrNum, carrier_group = carrier_group,
                             model = mdl_labs[3], expression = med_cleaned)
        ## report how many samples are plotted in the mediator panel for this run
        panel3_n <- plot_long_step2 |>
            dplyr::filter(carrier_group %in% c("E2+", "E4+")) |>
            nrow()
        message(sprintf(
            "[plotMedComp] mediator panel samples=%d mode=%s cluster=%s",
            panel3_n, med_mode_eff, as.character(mr$cluster)
        ))
        plot_long <- dplyr::bind_rows(plot_long_xy, plot_long_step2) |>
            dplyr::filter(carrier_group %in% c("E2+", "E4+")) |>
            dplyr::mutate(model = factor(as.character(model), levels = mdl_labs))
    }

    br_stats <- extract_gene_stats(dt_br, gene_id = out_gid, gene_col = "outcome_gene_id")
    mr_stats <- extract_gene_stats(dt_mr, gene_id = out_gid, gene_col = "outcome_gene_id")
    med_pre <- get_precomputed_med_stats(mr)
    med_recalc <- NULL
    if (med_mode_eff %in% c("recalc", "both")) {
        med_recalc <- get_recalc_med_stats(
            mr = mr, pd = pd, med_cache_dir = med_cache_dir,
            brnum_use = brnum_use
        )
    }
    if (med_mode_eff == "both") {
        med_labels <- c(
            sprintf("FDR: %s\nt: %s", format_fdr(med_pre$fdr), format_tstat(med_pre$t)),
            sprintf("FDR: %s\nt: %s", format_fdr(med_recalc$fdr), format_tstat(med_recalc$t))
        )
    } else {
        med_labels <- build_med_label(med_mode_eff, med_pre, med_recalc)
    }

    y_max <- max(plot_long$expression, na.rm = TRUE)
    y_min <- min(plot_long$expression, na.rm = TRUE)
    y_fixed <- y_max + pmax(0.12 * (y_max - y_min), 0.35)
    panel_tag_df <- dplyr::tibble(
        model = factor(mdl_labs, levels = mdl_labs),
        panel_tag = LETTERS[seq_along(mdl_labs)],
        carrier_group = "E2+",
        y_tag = Inf
    )
    anno <- dplyr::tibble(
        model = factor(mdl_labs, levels = mdl_labs),
        label = c(
            sprintf("FDR: %s\nt: %s", format_fdr(br_stats$fdr), format_tstat(br_stats$t)),
            sprintf("FDR: %s\nt: %s", format_fdr(mr_stats$fdr), format_tstat(mr_stats$t)),
            med_labels
        ),
        y = y_fixed
    )

    src_cfg <- pd$src$med_expr_src %||% NULL
    src_desc <- if (is.null(src_cfg)) {
        sprintf("Source context: %s", mcl)
    } else if (as.character(src_cfg$kind) == "erc_sce") {
        mcl <- paste0("ERC ",mcl)
        sprintf("Source context: %s cluster", mcl)
    } else if (as.character(src_cfg$kind) == "lc_sp_astro") {
        mcl <- sub("LC", "LC ", mcl) ## LCAstro -> LC Astro for better readability
        sprintf("Source context: %s spatial domain", mcl)
    } else {
        sprintf("Source context: %s", mcl)
    }
    strip_fills <- rep("#d1d1d1", length(mdl_labs))
    if (med_mode_eff == "both") {
        strip_fills[c(3, 4)] <- "#b0c8e2"
    } else if (length(strip_fills) >= 3) {
        strip_fills[3] <- "#b0c8e2"
    }

    out_rng <- range(plot_long$expression[plot_long$model %in% mdl_labs[1:2]], na.rm = TRUE)
    med_rng <- range(plot_long$expression[plot_long$model %in% mdl_labs[3:length(mdl_labs)]], na.rm = TRUE)
    ## add explicit top headroom per y-scale group for FDR/t annotations
    out_span <- diff(out_rng)
    med_span <- diff(med_rng)
    out_rng_pad <- c(out_rng[1], out_rng[2] + pmax(0.15 * out_span, 0.35))
    med_rng_pad <- c(med_rng[1], med_rng[2] + pmax(0.15 * med_span, 0.35))

    ggplot(plot_long, aes(x = carrier_group, y = expression, fill = carrier_group,
                        color = carrier_group)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA, linewidth = 0.3) +
        geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
        ggh4x::facet_wrap2(
            ~model,
            nrow = rows, scales = "free_y",
            strip = ggh4x::strip_themed(
                background_x = ggh4x::elem_list_rect(fill = strip_fills, colour = NA)
            )
        ) + ggh4x::facetted_pos_scales(
          y = c(replicate(2, scale_y_continuous(limits = out_rng_pad), simplify = FALSE),
                replicate(length(mdl_labs) - 2, scale_y_continuous(limits = med_rng_pad), simplify = FALSE)
          )
        ) +
        scale_fill_manual(values = APOE_carrier_colors) +
        scale_color_manual(values = APOE_carrier_colors_dark) +
        labs( title = sprintf("Outcome: ERC Oligo.3 %s | Mediator: %s %s", og, mcl, mediator_gene),
            ## subtitle = src_desc,
            x = "APOE Carrier Status",
            y = "Covariate-adjusted Expression (logCPM)",
            fill = "Carrier"
        ) +
            geom_text(
                data = anno,
                aes(x = 1.5, y = Inf, label = label),
                inherit.aes = FALSE,
                size = 3.5, vjust = 1.2
          ) +
        {
            geom_text(
                data = panel_tag_df,
                aes(x = carrier_group, y = y_tag, label = panel_tag),
                inherit.aes = FALSE,
                position = position_nudge(x = -0.7),
                size = 9,
                fontface = "bold",
                color = "black",
                hjust = 1.2,
                vjust = -0.65
            )
        } +
        coord_cartesian(clip = "off") +
        theme_bw(base_size = 13) +
        theme( legend.position = "none",
            ## title positioning:
            plot.title = element_text(margin = margin(b = 14)),
            strip.text = element_text(size = 10, face = "bold"),
            ## ticks:
            axis.text.x = element_text(size = 11, color = "grey40", face="bold"),
            axis.text.y = element_text(size = 11, color = "grey20"),
            ## axis titles:
            axis.title.x = element_text(size = 12,  margin = margin(t = 13)),
            axis.title.y = element_text(size = 12, color = "black"),
            panel.spacing.x = grid::unit(26, "pt"),
            panel.spacing.y = grid::unit(26, "pt"),
            panel.border = element_rect(color = "grey40", fill = NA, linewidth = 0.4),
            plot.margin = margin(t = 20, r = 12, b = 12, l = 12)
        ) # +  coord_cartesian(clip = "off")
}

#p <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", pd = pd, med_region = "LC")
#p <- plotMedComp(outcome_gene = "MBP", mediator_gene = "ABCA8", pd = pd, med_stats_mode = "both")
#print(p)
## load below the LC data first:
plc <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", pd = pd_lc,
                    med_region = "LC", med_stats_mode = "both")

print(plc)

```

### Comparison boxplots

LEFT panel: outcome expression before mediator adjustment.

MIDDLE panel: outcome expression after mediator adjustment.

RIGHT panel: mediator expression in its source context. The header strip is light blue to mark the different context.

How cleaningY(y, mod, P) works:
Fits y = mod %*% beta (full model)
Returns y - (effects from columns P+1 to end)
With P=4, preserves APOE columns (1-4), removes covariate effects
Input data needed:
 Outcome expression matrix (y_logcpm): precomputed as pd$mat$y_lcpm from edgeR::cpm(pd$obj$dge, log=TRUE)
 Mediator expression vector: From pd$mat$m_expr[med_gene_id, ] aligned to outcome samples
 Design matrices:
  Baseline: ~0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio
  Mediation: same + med_vec
  Sample metadata: pd$obj$dge$samples$APOE_syn to group E2.E2 vs E3.E4/E4.E4


```{r LC mediation data load and plot}
## loading LC mediation data
 lc_src_voom <- file.path(cdir, "22_Mediation", "LC-processed-data", "dgelE2E4_voomLmFit_object_MODEL2.qs2")
 lc_src_donor_pb <- file.path(cdir, "22_Mediation", "LC-processed-data", "shared_internal_ref",
                              "DGEList_pseudobulkcounts_by_donor_forGP_unfilt_020426.RDS")
 pd_lc <- loadMedData(
     file.path(cdir, "22_Mediation", "out-lc_astro"),
     med_expr_src = list(
         kind = "lc_sp_astro",
         path = lc_src_voom,
         meta = list(donor_pb_path = lc_src_donor_pb)
     )
 )
plc <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", pd = pd_lc,
                    med_region = "LC", med_stats_mode = "recalc")

print(plc)
```


```{r LC Astro mediator DGE stats using original tissue-section expression data, without donor-pseudobulk}
plc <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", pd = pd_lc,
                    med_region = "LC", med_stats_mode = "both")

print(plc)
```

```{r}
plc2row <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", pd = pd_lc,
                    med_region = "LC", med_stats_mode = "both", rows = 2)

print(plc2row)


```





```{r plots in PDFs for specific genes - ERC Astro case}
dnr=c('PLP1', 'MAG', 'MAL', 'MBP', 'SOX10', 'OPALIN')
upr=c('FOS', 'TLR2', 'STAT1', 'STAT4')
alldegs=c(dnr, upr)
medgenes=c('PLP1', 'MBP', 'OPALIN')
pdf_dir <- pd$pth$out
pdf_files <- file.path(pdf_dir, sprintf("mediator_%s.pdf", medgenes))

plot_one_mediator <- function(mediator_gene, outcome_genes, out_file) {
    grDevices::pdf(out_file, width = 11, height = 8.5, onefile = TRUE)
    on.exit(grDevices::dev.off(), add = TRUE)

    for (outcome_gene in outcome_genes) {
        p <- plotMedComp(outcome_gene = outcome_gene, mediator_gene = mediator_gene, pd = pd)
        print(p)
    }

    invisible(out_file)
}

pdf_out <- mapply(
    plot_one_mediator,
    mediator_gene = medgenes,
    out_file = pdf_files,
    MoreArgs = list(outcome_genes = alldegs),
    SIMPLIFY = FALSE
)

pdf_out

```

```{r load and plot LC mediation examples}



```
