---
title: "Plot mediation screening results"
## format: html
---

```{r}
library(edgeR)
library(SingleCellExperiment)
library(data.table)
library(ggplot2)
library(dplyr)
library(here)
ddir <- here("processed-data")
```



```{r setup}
library(qs2)
library(ggplot2)
library(tidyr)
## assumes 01_mediation_screen.R was already run for this scenario
cdir <- here("code")

loadMedData <- function(out_dir, src_expr = NULL) {
    if (!is.null(src_expr)) {
        stop("loadMedData(src_expr=...) is not implemented yet.")
    }

    pth <- list(
        out = out_dir,
        base_dir = file.path(out_dir, "baseline"),
        fn_md = file.path(out_dir, "med_degs.qs2"),
        fn_me = file.path(out_dir, "med_expr.qs2"),
        fn_dge = file.path(out_dir, "dge_base.qs2"),
        fn_mi = file.path(out_dir, "mediation_run_index.tsv"),
        fn_bi = file.path(out_dir, "baseline_index.tsv")
    )
    if (!file.exists(pth$fn_md)) stop("Missing `med_degs.qs2`: ", pth$fn_md)
    if (!file.exists(pth$fn_me)) stop("Missing `med_expr.qs2`: ", pth$fn_me)
    if (!file.exists(pth$fn_dge)) stop("Missing `dge_base.qs2`: ", pth$fn_dge)

    md <- data.table::as.data.table(qs_read(pth$fn_md))
    if (!is.data.frame(md) || nrow(md) < 1) {
        stop("`med_degs` is empty or invalid in: ", pth$fn_md)
    }
    me <- qs_read(pth$fn_me)
    if (!is.matrix(me) || nrow(me) < 1) {
        stop("`med_expr` is empty or invalid in: ", pth$fn_me)
    }
    dge <- qs_read(pth$fn_dge)

    mi <- if (file.exists(pth$fn_mi)) data.table::fread(pth$fn_mi) else NULL
    bi <- if (file.exists(pth$fn_bi)) data.table::fread(pth$fn_bi) else NULL

    bf <- NULL
    if (!is.null(bi) && nrow(bi) > 0 && "out_file" %in% names(bi)) {
        bi2 <- data.table::copy(bi)
        if ("n_donors" %in% names(bi2)) {
            data.table::setorderv(bi2, cols = "n_donors", order = -1L)
        }
        bf <- bi2$out_file[1]
    } else {
        bf <- list.files(pth$base_dir, pattern = "^baseline_[0-9a-f]+\\.tab\\.gz$", full.names = TRUE)[1]
    }
    if (is.na(bf) || !nzchar(bf) || !file.exists(bf)) {
        stop("Could not resolve baseline file in: ", pth$base_dir)
    }
    br <- data.table::fread(bf)
    if (!is.data.frame(br) || nrow(br) < 1) {
        stop("`baseline_results_dt` is empty in: ", bf)
    }

    if (is.null(dge$genes) || !("gene_name" %in% colnames(dge$genes))) {
        stop("`dge_base$genes` must contain `gene_name`.")
    }
    gmap <- data.table::data.table(
        gene_name = as.character(dge$genes$gene_name),
        gene_id = rownames(dge$genes)
    )
    brnum <- gsub("_Oligo\\.3$", "", colnames(dge))
    ap <- as.character(dge$samples$APOE_syn)
    grp <- ifelse(ap %in% c("E2.E2", "E2.E3"), "E2+",
        ifelse(ap %in% c("E3.E4", "E4.E4"), "E4+", "other")
    )

    cfg <- list(
        base_str = "0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio",
        n_apoe = 4L
    )
    cfg$med_str <- paste(cfg$base_str, "+ med_vec")
    cfg$base_f <- as.formula(paste("~", cfg$base_str))
    cfg$med_f <- as.formula(paste("~", cfg$med_str))

    pd <- list(
        pth = pth,
        dt = list(md = md, mi = mi, bi = bi, br = br),
        obj = list(dge = dge),
        mat = list(y_lcpm = edgeR::cpm(dge, log = TRUE), m_expr = me),
        key = list(gmap = gmap, brnum = brnum, grp = grp),
        cfg = cfg
    )
    pd
}


```

```{r load mediation data}
## loading ERC Astro mediation data
pd <- loadMedData(file.path(cdir, "22_Mediation", "out-erc_astro"))
```


```{r plot function and test}
APOE_carrier_colors_dark <- c(`E2+`="#3B918B", `E4+`="#B7522A")
APOE_carrier_colors <- c(`E2+`="#51B8B1", `E4+`="#D97D59")

resolve_mediator_row <- function(mg, md) {
    mg_cl <- NULL
    mg_nm <- mg
    if (grepl(":", mg, fixed = TRUE)) {
        pr <- strsplit(mg, ":", fixed = TRUE)[[1]]
        mg_cl <- pr[1]
        mg_nm <- pr[2]
    }
    rs <- md[
        gene_name == mg_nm | gene_id == mg_nm | med_gene_id == mg_nm
    ]
    if (!is.null(mg_cl)) {
        rs <- rs[cluster == mg_cl]
    }
    if (nrow(rs) == 0) {
        cands <- unique(md$gene_name)
        cls <- cands[agrep(mg_nm, cands, max.distance = 0.2)]
        stop("Mediator not found: ", mg, "\n",
            "Tried gene_name / gene_id / med_gene_id.\n",
            "Close matches: ", paste(head(cls, 6), collapse = ", "))
    }
    if (nrow(rs) > 1) {
        warning(sprintf(
            "Multiple matches for mediator gene '%s' in clusters: %s. Using last match; use 'cluster:gene' to disambiguate.",
            mg, paste(unique(rs$cluster), collapse = ", ")
        ))
    }
    rs[nrow(rs)]
}

plotMedComp <- function(outcome_gene, mediator_gene, pd, med_region = "ERC", n_apoe_cols = 4L) {
    if (is.null(pd) || !is.list(pd)) stop("`pd` must be a plotData list from loadMedData().")

    dge <- pd$obj$dge
    dt_md <- pd$dt$md
    dt_mi <- pd$dt$mi
    dt_br <- pd$dt$br
    m_expr <- pd$mat$m_expr
    y_lcpm <- pd$mat$y_lcpm
    gmap <- pd$key$gmap
    brnum <- pd$key$brnum
    grp <- pd$key$grp
    base_f <- pd$cfg$base_f
    med_f <- pd$cfg$med_f
    n_ap <- as.integer(n_apoe_cols)

    mr <- resolve_mediator_row(mediator_gene, dt_md)
    run_lb <- sprintf("%s|%s|%s", mr$cluster, mr$gene_id, mr$gene_name)
    f_lb <- gsub("\\|", "_", run_lb)
    it <- as.integer(mr$iter)
    mgid <- as.character(mr$med_gene_id)
    mcl <- as.character(mr$cluster)

    mf <- character(0)
    if (!is.null(dt_mi) && nrow(dt_mi) > 0 &&
        all(c("iter", "med_gene_id", "out_file") %in% names(dt_mi))) {
        idx <- dt_mi[iter == it & med_gene_id == mgid]
        if (nrow(idx) > 0) {
            mf <- idx$out_file[1]
        }
    }
    if (length(mf) < 1 || !file.exists(mf)) {
        mf <- file.path(pd$pth$out, sprintf("%03d_%s.tab.gz", it, f_lb))
    }
    if (!file.exists(mf)) {
        cands <- list.files(pd$pth$out, pattern = sprintf("_%s\\.tab\\.gz$", f_lb), full.names = TRUE)
        if (length(cands) > 1) {
            finfo <- file.info(cands)
            cands <- cands[order(finfo$mtime, decreasing = TRUE)]
        }
        mf <- cands[1]
    }
    if (is.na(mf) || !nzchar(mf) || !file.exists(mf)) {
        stop("Mediation results file not found for mediator: ", mediator_gene)
    }
    dt_mr <- data.table::fread(mf)

    if (!mgid %in% rownames(m_expr)) {
        stop("Mediator expression not found in `pd$mat$m_expr` for: ", mgid)
    }

    og <- outcome_gene
    ghit <- gmap[gene_name == og]
    if (nrow(ghit) < 1) {
        stop("Outcome gene not found in dge_base: ", outcome_gene)
    }
    out_gid <- ghit$gene_id[1]

    y_mat <- matrix(y_lcpm[out_gid, ], nrow = 1)
    rownames(y_mat) <- out_gid
    colnames(y_mat) <- colnames(y_lcpm)

    med_vec <- as.numeric(m_expr[mgid, brnum])
    if (anyNA(med_vec)) {
        stop("Found NA values in mediator vector for: ", mgid)
    }

    mod_b <- model.matrix(base_f, data = dge$samples)
    smp <- dge$samples
    smp$med_vec <- med_vec
    mod_m <- model.matrix(med_f, data = smp)

    yb <- jaffelab::cleaningY(y_mat, mod_b, P = n_ap)
    ym <- jaffelab::cleaningY(y_mat, mod_m, P = n_ap)

    plot_df <- data.frame(
        BrNum = brnum,
        carrier_group = grp,
        baseline_cleaned = as.numeric(yb[1, ]),
        med_cleaned = as.numeric(ym[1, ])
    )
    plot_df <- plot_df[plot_df$carrier_group %in% c("E2+", "E4+"), ]

    mdl_labs <- c(
        sprintf("Baseline\n(no %s)", mediator_gene),
        sprintf("After %s adjustment", mediator_gene)
    )

    plot_long <- plot_df |>
        tidyr::pivot_longer(
            cols = c(baseline_cleaned, med_cleaned),
            names_to = "model",
            values_to = "expression"
        ) |>
        dplyr::mutate(model = factor(model,
             levels = c("baseline_cleaned", "med_cleaned"),
             labels = mdl_labs)
        )
    anno <- NULL
    if (!is.null(dt_br) && !is.null(dt_mr)) {
        br_row <- dt_br[outcome_gene_id == out_gid, , drop = FALSE]
        mr_row <- dt_mr[outcome_gene_id == out_gid, , drop = FALSE]
        br_row <- if (nrow(br_row) > 0) br_row[1, ] else NULL
        mr_row <- if (nrow(mr_row) > 0) mr_row[1, ] else NULL

        fdr_b <- if (!is.null(br_row) && "adj.P.Val" %in% names(br_row)) br_row$adj.P.Val else NA_real_
        t_b <- if (!is.null(br_row) && "t" %in% names(br_row)) br_row$t else NA_real_
        fdr_m <- if (!is.null(mr_row) && "adj.P.Val" %in% names(mr_row)) mr_row$adj.P.Val else NA_real_
        t_m <- if (!is.null(mr_row) && "t" %in% names(mr_row)) mr_row$t else NA_real_

        y_max <- max(plot_long$expression, na.rm = TRUE)
        y_min <- min(plot_long$expression, na.rm = TRUE)
        y_fixed <- y_max + pmax(0.08 * (y_max - y_min), 0.2)

        format_fdr <- function(fdr) {
            if (is.na(fdr)) return("NA")
            if (fdr >= 1e-4) {
                sprintf("%.5f", fdr)
            } else {
                sprintf("%.2e", fdr)
            }
        }
        anno <- dplyr::tibble(
            model = factor(mdl_labs, levels = mdl_labs),
            label = c(
                sprintf("FDR: %s\nt: %.2f", format_fdr(fdr_b), t_b),
                sprintf("FDR: %s\nt: %.2f", format_fdr(fdr_m), t_m)
            ),
            y = y_fixed
        )
    }
    subtitle <- fifelse(med_region == "ERC",
        sprintf("Mediator: %s expression from %s %s cluster", mediator_gene, med_region, mcl),
        ## LC:
        sprintf("Mediator: %s expression from %s spatial domain", mediator_gene, mcl)
    )
    ggplot(plot_long, aes(x = carrier_group, y = expression, fill = carrier_group,
                        color = carrier_group)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA) +
        geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
        facet_wrap(~model) +
        scale_fill_manual(values = APOE_carrier_colors) +
        scale_color_manual(values = APOE_carrier_colors_dark) +
        labs( title = sprintf("%s (%s) Expression: E2+ vs E4+ Carriers",
                            og, out_gid),
            subtitle = subtitle,
            x = "APOE Carrier Status",
            y = "Covariate-adjusted Expression (logCPM)",
            fill = "Carrier"
        ) +
        {
            if (!is.null(anno)) {
                ggplot2::geom_text(
                    data = anno,
                    ggplot2::aes(x = 1.5, y = y, label = label),
                    inherit.aes = FALSE,
                    size = 3.5
                )
            }
        } +
        theme_bw(base_size = 14) +
        theme( legend.position = "none",
            strip.text = element_text(size = 12, face = "bold")
        ) # +  coord_cartesian(clip = "off")
}

#p <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", pd = pd, med_region = "LC")
p <- plotMedComp(outcome_gene = "MBP", mediator_gene = "ABCA8", pd = pd)
print(p)
```

### Comparison boxplots BEFORE vs AFTER mediation

LEFT panel (Baseline): E2 vs E4 boxplots showing Oligo.3 gene expression before adding the mediator (Astro PLP1) as a covariate. The expression values are "cleaned" using jaffelab::cleaningY to remove effects of nuisance covariates (Sex, Age, Anc_Afr, pseudo_expr_chrM_ratio) while preserving the APOE effect.

RIGHT panel (After mediation): E2 vs E4 boxplots showing the same outcome gene expression, but now cleaned to remove both nuisance covariates AND the mediator gene expression. If the mediator explains part of the APOE→Oligo.3 effect, the E2 vs E4 difference should shrink.

How cleaningY(y, mod, P) works:
Fits y = mod %*% beta (full model)
Returns y - (effects from columns P+1 to end)
With P=4, preserves APOE columns (1-4), removes covariate effects
Input data needed:
 Outcome expression matrix (y_logcpm): precomputed as pd$mat$y_lcpm from edgeR::cpm(pd$obj$dge, log=TRUE)
 Mediator expression vector: From pd$mat$m_expr[med_gene_id, ] aligned to outcome samples
 Design matrices:
  Baseline: ~0 + APOE_syn + Sex + Age + Anc_Afr + pseudo_expr_chrM_ratio
  Mediation: same + med_vec
  Sample metadata: pd$obj$dge$samples$APOE_syn to group E2.E2 vs E3.E4/E4.E4


```{r LC mediation data load and plot}
## loading LC mediation data
 pd_lc <- loadMedData(file.path(cdir, "22_Mediation", "out-lc_astro"))
 plc <- plotMedComp(outcome_gene = "MMD2", mediator_gene = "SLC1A3", pd = pd_lc, med_region = "LC")
print(plc)
```








```{r plots in PDFs for specific genes - ERC Astro case}
dnr=c('PLP1', 'MAG', 'MAL', 'MBP', 'SOX10', 'OPALIN')
upr=c('FOS', 'TLR2', 'STAT1', 'STAT4')
alldegs=c(dnr, upr)
medgenes=c('PLP1', 'MBP', 'OPALIN')
pdf_dir <- pd$pth$out
pdf_files <- file.path(pdf_dir, sprintf("mediator_%s.pdf", medgenes))

plot_one_mediator <- function(mediator_gene, outcome_genes, out_file) {
    grDevices::pdf(out_file, width = 11, height = 8.5, onefile = TRUE)
    on.exit(grDevices::dev.off(), add = TRUE)

    for (outcome_gene in outcome_genes) {
        p <- plotMedComp(outcome_gene = outcome_gene, mediator_gene = mediator_gene, pd = pd)
        print(p)
    }

    invisible(out_file)
}

pdf_out <- mapply(
    plot_one_mediator,
    mediator_gene = medgenes,
    out_file = pdf_files,
    MoreArgs = list(outcome_genes = alldegs),
    SIMPLIFY = FALSE
)

pdf_out

```

```{r load and plot LC mediation examples}



```
